\title{SVD-Compressed Cross Sections in Monte
Carlo Neutron Transport\\
\large An implementation-led benchmark with positive, mixed,
and negative results; with a partition-of-unity fix for
off-library temperature interpolation}

\author{F.~Rodrigues}
\date{\today}
\maketitle

\begin{abstract}
Continuous-energy Monte Carlo transport codes spend a large
fraction of wall-clock time looking up microscopic cross
sections in pointwise tables indexed by energy and temperature.
We set out to test whether a truncated singular value
decomposition (SVD) of $\log_{10} \sigma(E, T)$ is a
defensible alternative to the pointwise table, and whether a
hybrid that couples SVD with Windowed Multipole (WMP) over the
resolved resonance region offers a compelling
memory--throughput profile. We built a pure-Rust transport
engine that implements a pointwise-table provider, an SVD
provider, an ACE+WMP provider that mirrors the
industry low-memory baseline, and a WMP+SVD hybrid provider
behind a common interface, validated the engine against
OpenMC~$0.15.3$ (closing an initial $78$--$127$~pcm offset to
$|\Delta| \le 51$~pcm across all four CPU/GPU providers after
tracing three transport-level bugs), and measured the
representations head-to-head on a $9$-nuclide thermal PWR
pin cell and on the Godiva (HEU-MET-FAST-001) fast benchmark
\emph{at both on-library and off-library operating temperatures}.
The findings are a mixture of positive, mixed, and negative,
and we report them plainly.

\textbf{Positive.}
(i)~The singular spectrum of ENDF/B-VII.1 $\log\sigma$
matrices decays rapidly. $43$ of $47$ non-redundant $^{235}$U
reaction channels are rank-one to machine precision.
Rank-$6$ reconstructs the full energy grid at machine precision
at $44$~ns per lookup against $91$~ns for a binary-search
pointwise table (kernel-level $2.0\times$).
(ii)~In-engine, SVD rank-$5$ agrees with the same engine's
pointwise table to $13$--$47$~pcm on PWR pin cell $k_\infty$
across two hardware tiers; all four in-engine providers
(CPU table/SVD, GPU pointwise/SVD) agree with OpenMC to
$|\Delta| \le 51$~pcm at large-L3 ten-seed statistics. On
Godiva the four providers cluster $325$--$362$~pcm above the
ICSBEP experimental value, which is the known ENDF/B-VII.1
Godiva bias; the engine reproduces the library-level bias
and does not add to it.
(iii)~End-to-end CPU SVD rank-$5$ runs
$1.37\times$--$1.90\times$ faster than the in-engine
pointwise table on PWR pin cell, hardware-dependent; on
Godiva CPU SVD is $1.43\times$ faster than the table on the
large-L3 system.
(iv)~\emph{Off-library result, headline.}
At a realistic operating temperature $+150$~K above every
nuclide's library column (fuel $900\!\to\!1050$~K,
moderator $600\!\to\!750$~K), with OpenMC-style stochastic
$T$-interpolation enabled on the pointwise and hybrid paths,
SVD with partition-of-unity $3$-point Ducru weights
(Sec.~\ref{sec:ducru_puoi}) reproduces $k_\infty$ within
$213$~pcm of the ACE+WMP industry baseline ($\sim\!3.6\sigma$
at ten-seed, $50\,000$-particle statistics) while consuming
$22\%$ less memory and running at speed parity
(within $0.07\%$). This closes a $2094$~pcm kernel-
approximation bias measured with a naive $2$-point Ducru
variant by a factor of $\approx\!10\times$
(Sec.~\ref{sec:off_library}).

\textbf{Mixed.}
(v)~The trade-off on GPU is opposite. GPU pointwise is
$1.30\times$ \emph{faster} than GPU SVD rank-$5$ on PWR pin
cell because four clad nuclides lack an MT$=4$ block and
their total inelastic must be synthesised from $13$
discrete-level SVD kernels per lookup. SVD is a CPU win on
this geometry and a GPU draw-to-loss.
(vi)~The hybrid SVD+WMP provider reproduces $k_\infty$ to
$86$~pcm of its own SVD baseline, but runs $2.06\times$
slower than CPU SVD because the Humlicek W4 Faddeeva
evaluation is $\sim\!15\times$ more expensive per lookup than
the SVD FMA. WMP buys memory (see next point) at the cost of
throughput.
(vii)~Per-reaction rank tuning is a modest but non-trivial
memory win. Reducing the $41$ discrete inelastic levels
(MT$=51$--$91$) from the global rank-$5$ to rank-$1$
(reviewer suggestion at PHYSOR workshop; a level's shape is
weakly $T$-dependent so one basis vector suffices) removes
$\approx\!70\%$ of the engine's SVD working set on Godiva
with no measurable $k_\text{eff}$ shift. The same mechanism
is what enables the off-library SVD memory win above.

\textbf{Negative.}
(viii)~The WMP representation packs the resolved-resonance
data of our nine nuclides into $1.37$~MB against $177.7$~MB
for a hypothetical four-channel ($\sigma_\text{el},
\sigma_\text{t}, \sigma_\text{f}, \sigma_\text{cap}$),
six-temperature pointwise table on the same nuclide set
($132.9\times$). That is a density ratio between two data
layouts, not the memory reduction an actual Monte Carlo
engine sees. Our engine carries SVD kernels for the discrete
inelastic levels ($41$ per U nuclide, $10$--$13$ per Zr),
for total inelastic, for $(n,2n)$ and $(n,3n)$, and for the
URR residue total; none of these are covered by the public
MIT WMP library. The measured in-engine reduction from
hybrid over full-SVD is $1.06\times$; in absolute terms the
hybrid engine carries $5\times$ \emph{more} memory than the
pointwise-table baseline. The two numbers are both real;
they measure different things, and the engine-scale one is
the one that matters for a reader who plans to build a
transport solver.
(ix)~A straight port of Ducru's kernel reconstruction to
off-library $T$ interpolation is not a drop-in. The raw
$2017$ Eq.~$31$ weights~\cite{ducru2017} are L2-optimal in
the Doppler-kernel norm but are \emph{not} a partition of
unity ($\sum_j w_j \neq 1$ in general), which imprints a
log-space gain error on resonance peak heights when applied
to V$^\top$ columns of a $\log\sigma$ SVD. A naive $2$-point
bracketing variant leaves a $\sim\!2100$~pcm bias on PWR
$k_\infty$ at off-library $T$ (Sec.~\ref{sec:off_library_2t}).
A simplex-projected QP variant (clip negative weights,
renormalize to $\sum w = 1$) is \emph{worse}, not better,
because it is L2-optimal in weight space rather than in
kernel space; measured on U-$238$~MT$=102$ it raises the
held-out error from $1.5\%$ to $5\%$. The partition-of-unity
$3$-point Ducru that delivered the (iv) headline is the
narrowest honest fix; a fully kernel-consistent QP~\cite{ducru2017}
remains future work.

\textbf{Scope.}
Measurements are for a nine-nuclide thermal pin cell and a
three-nuclide fast sphere, on two consumer systems with
different CPU L3 size and GPU class. Full-core behaviour
($\gtrsim\!200$ nuclides, minor actinides, fission products,
multiple S($\alpha$,$\beta$) libraries) is not evaluated here,
and the compression gap on engine-level memory would widen or
shrink with the reaction-channel mix there. Off-library
temperature measurements use a global $+150$~K operating
offset; coupled fuel-temperature/moderator-temperature
feedback loops are not modelled.  We report the measurements
we made and flag what they do not cover.

The $132.9\times$ representation-byte ratio and the
$1.06\times$ engine-level reduction are both measured,
both reproducible, and measure different things; we make
that distinction explicit throughout the paper so a reader
who plans to build a transport engine on top of this work
can calibrate expectations.
\end{abstract}

\noindent\textbf{Keywords:} Monte Carlo, cross-section
compression, singular value decomposition, windowed multipole,
Doppler broadening, partition-of-unity interpolation,
nuclear data, GPU transport, benchmark, trade-offs.

\tableofcontents

