\section{Conclusion}

Truncated SVD of $\log_{10}\sigma(E,T)$ is a viable
drop-in for pointwise tables in continuous-energy Monte
Carlo neutron transport, with the following constraints
satisfied: rank~$5$ on the smooth channels, rank~$1$ on
the MT$=51$--$91$ discrete-level cascade,
three-point unity-normalized Ducru weights for off-library
temperature interpolation, and channel-consistent
stochastic $T$-picking. Validation against
ICSBEP HEU-MET-FAST-$001$ lands at $+79 \pm 38$~pcm,
inside the experimental $\pm100$~pcm uncertainty band;
on a $9$-nuclide PWR pin cell, agreement with the engine's
own pointwise table is $5$--$47$~pcm and with OpenMC
$0.15.3$ on the same library is $|\Delta| \le 51$~pcm at
ten-seed statistics.

\subsection{When SVD pays}
\label{sec:when}

\paragraph{Off-library temperatures.}
This is the regime where SVD's measurement story is
unambiguously positive. At a uniform $+150$~K offset on
PWR, SVD with three-point unity Ducru is $22\%$ smaller
than ACE+WMP in memory ($156$ vs $206$~MB), matches
its CPU throughput within $0.07\%$, and trails its
$k_\infty$ by $213$~pcm
($\sim\!3.6\sigma_\text{seed}$, Sec.~\ref{sec:off_library_3t}).
The pointwise table doubles its memory off-library because
stochastic $T$-picking requires both bracketing library
columns to be resident; SVD reconstructs the off-library
$\sigma$ from a single basis with three weights and does
not pay this cost.

\paragraph{High rank-$1$ reaction-channel mix.}
$43$ of $47$ U-$235$ reaction channels are rank-$1$ to
machine precision (Sec.~\ref{sec:spectrum}); the
weakly-$T$-dependent MT$=51$--$91$ discrete levels are
each rank-$1$ to machine precision across all seven
ENDF/B-VII.1 library columns. Where the reaction-channel
mix matches this profile, the rank budget is small and the
in-engine memory penalty for SVD is correspondingly
small.

\paragraph{Multi-temperature libraries.}
With $T = 6$ library columns (the standard
depletion-feedback multi-temp HDF5 set), the pointwise
baseline grows linearly in $T$ to $\sim\!600$~MB on the
$9$-nuclide set while pure SVD and hybrid stay at
$\sim\!519$~MB; the on-library memory ordering inverts
in SVD's favour. The measurements in this paper use
$T = 1$, so this is a projection, not a measured win.

\subsection{Where SVD does not pay}
\label{sec:where_not}

\paragraph{On-library, $T = 1$, full-physics deployment.}
At every rank measured ($1$, $2$, $3$, $5$, $7$), SVD
in-engine memory is monotonically larger than the
pointwise table at on-library temperatures
(Sec.~\ref{sec:memprec}). The basis stores $r$ values per
energy point; the table stores one. CPU throughput is
favourable ($1.37\times$--$1.90\times$ vs the in-engine
pointwise table on PWR), but on memory there is no
on-library rank that wins.

\paragraph{GPU.}
The earlier headline that GPU SVD was $1.30\times$ slower
than GPU pointwise on PWR has been retired. The cause was
the level-sum walk for nuclides whose ENDF/B-VII.1
evaluation omits MT$=4$ (Zr-$90$/$91$/$92$/$94$, U-$238$).
A per-warp $(n, E_\text{idx})$ shared-memory cache, named
in the previous draft as the obvious fix, was implemented
and falsified ($1.003\times$ measured at five seeds; energy-
bin sort coherence too low; preserved as a cautionary
result in §\ref{sec:threats}). The actual fix lives at
HDF5 load: synthesise MT$=4$ from the discrete-level sum
and tabulate the per-level CDF on a log-decimated grid.
Both together collapse the kernel hot path to a single
rank-$5$ \texttt{svd\_reconstruct} plus a binary CDF
search. On RTX A$1000$ the synth$+$CDF row at three seeds
$\times$ $60$ batches $\times$ $50\,000$ particles runs
$1.0099\times$ \emph{faster} than GPU pointwise on the same
geometry ($19\,473$ vs $19\,664$~ns/p), with $k_\infty$
agreeing to $+16$~pcm (well inside SEM).

\paragraph{Hybrid SVD+WMP throughput.}
The hybrid is $2.06\times$ slower than CPU SVD because
the Faddeeva evaluation inside each WMP window is
$\sim\!15\times$ more expensive than the SVD FMA. The
hybrid is shipped as a correctness deliverable, not as
a throughput improvement; it is the right path only when
WMP-grade resonance accuracy is required and the
$2\times$ slowdown is acceptable.

\subsection{Caveats and limits}
\label{sec:caveats}

\paragraph{Engine-scale memory $\neq$ representation-byte
ratio.}
The $132.9\times$ representation-byte ratio of
WMP$+$smooth-SVD against a hypothetical four-channel
multi-temperature pointwise table is correct
per-nuclide-per-channel
(Table~\ref{tab:hybrid_mem}), but the in-engine working
set on PWR is $4.8$--$5.2\times$ \emph{larger} than the
pointwise baseline at $T = 1$, because the discrete-level
SVD kernels (which WMP does not cover) dominate the
engine memory. The smooth-only-rebuild reduces this by
$1.06\times$ ($519.0 \to 487.6$~MB). Quoting a single
compression ratio without scope conflates these two
metrics; both are reported here so the comparison is
unambiguous.

\paragraph{Rank-$1$ on PWR collapses.}
Rank-$1$ SVD reconstructs U-$238$ resonance self-shielding
to a single basis vector and produces $k_\infty = 0.871$
on the on-library PWR pin cell ($\sim\!46\,000$~pcm
low). Rank-$2$ overshoots ($1.409$, $\sim\!8\,000$~pcm
high). Rank-$3$ is the lowest deployable rank
on resonance-dominated thermal systems; rank-$5$ is the
production floor.

\paragraph{Library-bias not investigated.}
Both the engine and OpenMC sit inside the ICSBEP
$\pm 100$~pcm uncertainty band on Godiva, from opposite
sides; the $+178$~pcm cross-code spread between them is
larger than either's offset from experiment. A fully
independent reference (MCNP, Serpent) would localise the
spread to one code or to the underlying ENDF/B-VII.1
evaluation. This is not investigated here.

\paragraph{S($\alpha,\beta$) coverage limited to H/H$_2$O.}
D$_2$O, graphite, ZrH and other thermal libraries are not
validated against an independent reference at this stage.

\paragraph{Charged-particle channels absorbed into capture.}
Channels MT$=103$, $107$, and first-chance fission
(MT$=19$--$21$) are folded into the capture macroscopic.
$k_\infty$ is correct under this convention; MT-level
reaction-rate tallies are not.

\paragraph{Single-seed hybrid statistics.}
The end-to-end Hybrid SVD+WMP $k_\infty = 1.32795 \pm
0.00103$ ($-26$~pcm vs ACE+WMP) is single-seed only at
the matched $150 \times 50\,000$ statistics; the
ten-seed redo is not in this paper.

\paragraph{CP/PARAFAC of $\sigma(E, T, \ell)$ — measured,
deferred.}
\label{sec:cp_probe}
The per-level rank-$1$ SVDs we currently store ignore
cross-$\ell$ correlation. We probed CP/PARAFAC --- a
joint rank-$R$ decomposition $\sigma_\ell(E,T) \approx
\sum_{r=1}^R a_r(E)\, b_r(T)\, c_r(\ell)$ --- via greedy
rank-$1$ deflation (\texttt{src/cp\_decompose.rs}) and
report results below at $200$ log-spaced energy points
across the six ENDF/B-VII.1 library temperatures
(\texttt{outputs/cp\_analysis.txt}).

\begin{tabular}{lrrrrr}
\toprule
nuclide & $N_\ell$ & rank-$3$ & rank-$5$ & rank-$8$ & mem ratio \\
        &          & rel.\ L2 & rel.\ L2 & rel.\ L2 & at rank-$8$ \\
\midrule
Zr-$90$ & $10$ & $7.0\%$ & $1.4\%$  & $0.26\%$  & $0.84\times$ \\
Zr-$91$ & $10$ & $3.6\%$ & $0.78\%$ & $0.046\%$ & $0.84\times$ \\
Zr-$92$ & $10$ & $8.6\%$ & $2.9\%$  & $0.32\%$  & $0.84\times$ \\
Zr-$94$ & $13$ & $9.8\%$ & $3.5\%$  & $0.82\%$  & $0.66\times$ \\
U-$238$ & $41$ & $6.0\%$ & $1.9\%$  & $0.45\%$  & $\mathbf{0.23\times}$ \\
\bottomrule
\end{tabular}

\noindent
The pattern is the predictable one: CP wins as $N_\ell$
grows because the cost of the $\ell$-mode factor
($N_\ell$ doubles per rank) is amortised against the
constant overhead of the $E$-mode and $T$-mode factors.
For U-$238$ at rank-$8$, CP reaches sub-pcm RMSE while
storing $4.3\times$ less than the $41$ per-level rank-$1$
SVDs. For Zr clad with $10$--$13$ levels, CP only barely
breaks even on memory at the precision we care about.

We do not deploy CP because synthesis $+$ CDF already
crossed parity with GPU pointwise at the per-particle
cost, and CP loses that comparison: a rank-$R$ CP
reconstruction at one $(E_\text{idx}, T)$ costs
$R(N_\ell + 2)$ FMAs to materialise all levels at once
($R = 5, N_\ell = 41 \to 215$~FMAs) versus the
synthesis-plus-CDF hot path's one rank-$5$
\texttt{svd\_reconstruct} plus one binary CDF search
($\sim 18$ FMA-equivalents). CP becomes the right pick
for full-actinide libraries with many high-level
nuclides (Pu isotopes, minor actinides), and for
analyses where all $13$--$41$ level cross sections are
queried in batch (sensitivity tallies, depletion
chains). Both are out of scope here.

\subsection{Bottom line}

SVD compression of ENDF/B-VII.1 cross sections is the
right choice for off-library, multi-temperature, or
rank-$1$-channel-heavy operating regimes on CPU; the
pointwise table remains the right choice for on-library,
single-temperature, GPU-deployed, or memory-constrained
configurations. The signed three-point unity Ducru
scheme is the production default at rank~$5$; the
kernel-Gram QP wins at low rank but loses at the
production rank. Three structural properties of
ENDF/B-VII.1 --- the rank-$1$ majority of U-$235$ reaction
channels, the weakly-$T$-dependent discrete-level
cascade, and the rank deficiency of $\log\sigma(E,T)$ at
fixed $E$ --- are real and exploitable; the raw Ducru
weights are not a partition of unity and must be
normalized before being applied to a $\log\sigma$ SVD.
