\section{GPU transport}
\label{sec:gpu}

The event-based CUDA solver~\cite{tramm2024} implements both
providers inside the same kernel; a runtime
\texttt{-{}-force-svd} flag clears the uploaded pointwise
tables so the SVD path reconstructs every reaction channel
from the basis.

\paragraph{The original gap, and the wrong fix.}
Earlier drafts of this paper reported GPU pointwise as
$1.30\times$ faster than GPU SVD on PWR ($5\,967$ vs $7\,754$~ns/p
on the large-L3 system) and flagged a per-warp shared-memory cache
keyed on $(n, E_\text{idx})$ as the engineering item that would
close it. The cause was geometry-specific: four clad nuclides
(Zr-$90$, $91$, $92$, $94$) do not carry an MT$=4$ total-inelastic
block in the ENDF/B-VII.1 HDF5; nor does U-$238$. The macroscopic
inelastic XS was therefore synthesised at lookup time by summing
$13$--$41$ discrete-level SVD kernels per step, which on GPU
dominated the SVD-path runtime.

We implemented the per-warp cache and measured it. With careful
threshold-gating (a first attempt produced a $+730$~pcm $k_\infty$
bias because lanes straddling a level threshold inherited the
leader's gate), it preserved $k_\infty$ within SEM
($\Delta k_\infty = -18.5$~pcm, three-SEM gate $208.8$~pcm) but
gave a throughput ratio of only $1.003\times$
($23\,508 \to 23\,434$~ns/p). The energy-bin sort granularity
($256$ log bins across $\sim\!17$ decades, $\sim\!6\%$ bin width) is
too coarse to make warp-level $(n, E_\text{idx})$ uniformity common,
so the cache rarely activates. Diagnosis preserved here as a
falsified hypothesis (and in §\ref{sec:threats}).

\paragraph{The right fix: synthesise MT$=4$ at load + tabulate the
per-level CDF.}
Both halves of the redundant work --- the macroscopic XS sum and
the level-selection walk --- are eliminated upstream of the kernel.
At HDF5 load, for every nuclide whose evaluation omits MT$=4$ we
sum the raw per-level cross sections across MT$=51$--$91$ on the
union grid and SVD-decompose the result, slotting the kernel into
the regular MT$=4$ position. Independently we precompute the
per-level CDF
\(F_\ell(E,T) = \sum_{l \leq \ell} \sigma_l(E,T) / \sigma_4(E,T)\)
on a $200$-point log-decimated energy grid (the CDF is smooth in $E$
because resonance peaks cancel in the $\sigma_\ell/\Sigma\sigma$
ratio; full-grid storage costs $357$~MB for U-$238$, decimated
$384$~KB at sub-pcm reconstruction error).

The kernel hot path then runs one rank-$5$ \texttt{svd\_reconstruct}
on the macroscopic XS (the synthesised MT$=4$) and one binary search
in the CDF to sample the level. The $13$--$41$-deep walk vanishes.
On a small-L3 RTX A$1000$, three seeds $\times$ $60$ batches
$\times$ $50\,000$ particles, $\mathtt{-{}-force\text{-}svd}$:

\begin{tabular}{lrr}
\toprule
provider & ns/p & $k_\infty$ \\
\midrule
GPU pointwise (reference)   & $19\,664 \pm 30$ & $1.32758 \pm 0.00063$ \\
GPU SVD pre-synth, cache off & $23\,508 \pm 58$ & $1.32759 \pm 0.00083$ \\
GPU SVD synth-only          & $19\,848 \pm 31$ & $1.32776 \pm 0.00083$ \\
GPU SVD synth $+$ CDF       & $19\,473 \pm 15$ & $1.32774 \pm 0.00043$ \\
\bottomrule
\end{tabular}

\noindent
Original ratio $1.30\times$ slower
$\to$ $1.0094\times$ slower (synth)
$\to$ $0.99\times$ \emph{faster} than GPU pointwise on the same
geometry. Physics fidelity holds: the synth$+$CDF row sits
$+16$~pcm above GPU pointwise, well inside combined SEM. The
per-warp cache, having been falsified, is removed in this commit;
the kernel is back to its pre-cache shape modulo the synth and CDF
plumbing.

\paragraph{Fidelity parity at large-L3 statistics.}
Post-correction GPU rows: GPU pointwise $k_\infty =
1.32821 \pm 0.00033$ and GPU SVD $k_\infty = 1.32762 \pm
0.00034$. The two GPU paths agree to $59$~pcm
($\approx 1.3\,\mathrm{SEM}_\text{combined}$), the two CPU
paths agree to $13$~pcm, and all four providers agree on
$k_\infty$ within $|\Delta_\text{OpenMC}| \le 51$~pcm with
the corrected stochastic-bin sampler. The GPU pointwise
row is the high point of the spread ($+51$~pcm vs OpenMC)
and remains within $3\,\mathrm{SEM}_{10}$; the other three
sit in the $-8$ to $-25$~pcm band.

%% ============================================================
