\section{GPU WMP evaluator}
\label{sec:gpu_wmp}

WMP evaluation is compute-dense (Faddeeva + broadened
curvefit + pole sum) with a small working set (few KB of
pole data per nuclide), which matches GPU arithmetic
pipelines well. We ported the CPU evaluator to CUDA as
\texttt{\_\_device\_\_} helpers in
\texttt{gpu/cuda/transport.cu} and added a
\texttt{wmp\_test\_eval} kernel plus a host validator binary
\texttt{rust\_prototype/src/bin/gpu\_wmp\_validate.rs}.

\paragraph{Bit-exactness.}
CPU-vs-GPU over twelve U-$238$ test energies at $T = 293.6$~K:
maximum relative discrepancy $5.3 \cdot 10^{-14}$ on
absorption and $2.0 \cdot 10^{-13}$ on fission --- double
precision round-off only.

\paragraph{Throughput on RTX~$3080$.}
On a $10^6$-energy log-spaced grid inside U-$238$'s RRR:

\begin{center}
\begin{tabular}{@{}lrr@{}}
\toprule
Implementation & ns/lookup & Relative \\
\midrule
CPU single-thread (Ryzen~$9800$X3D) & $66.8$ & $1.00\times$ \\
GPU (RTX~$3080$, saturated)         & $9.5$  & $0.14\times$ \\
\bottomrule
\end{tabular}
\end{center}

$7.0\times$ per-lookup GPU-vs-CPU.

\paragraph{End-to-end GPU hybrid transport.}
The validated evaluator is wired into the event-based
\texttt{transport\_persistent} kernel: for each nuclide that
carries WMP data, a bitmap \texttt{has\_wmp[i]} and the
corresponding $E_{\min}^{\mathrm{WMP}}, E_{\max}^{\mathrm{WMP}}$
are checked inline; if the incident energy falls in the
resolved-resonance window, the SVD / pointwise lookup for
elastic, fission and capture is replaced by \texttt{wmp\_eval}
at the nuclide's target temperature, the total XS is
recomputed from partials, and URR sampling is suppressed
(resonances are already explicit through the poles).
Inelastic, $(n,2n)$ and $(n,3n)$ remain on the SVD path
because WMP does not cover those channels.

To keep this practical on GPU the recursive conjugate fold
in the Humlicek Faddeeva implementation is unrolled into a
flag-plus-sign-flip form, because recursive \verb|__device__|
calls spill to local memory under the register pressure of
\texttt{\_\_launch\_bounds\_\_(256,\,2)} and trigger
CUDA\_ERROR\_ILLEGAL\_ADDRESS on deep stacks. The rewritten
evaluator is still bit-exact against the CPU reference at
$5\cdot10^{-14}$.

On the PWR pin cell, $2$ of $9$ nuclides are covered by WMP
(U-$235$ and U-$238$ in the fuel; the released ENDF/B-VII.1
WMP library does not ship U-$234$, O-$16$, H-$1$ or Zircaloy
isotopes in a form the reader accepts). For those two
nuclides the GPU hybrid path exercises \texttt{wmp\_eval}
throughout each nuclide's resolved-resonance window and falls
back to the SVD / pointwise path elsewhere; covered nuclides
therefore receive the explicit multipole treatment inside
the RRR while uncovered ones keep the existing high-fidelity
SVD path. $k_\infty$ and throughput numbers from the large-L3
benchmark are in Section~\ref{sec:pwr}.

%% ============================================================
