\section{Implementation}
\label{sec:impl}

The engine is pure Rust with no C dependency; nuclear data
reads directly from OpenMC HDF5 files via \texttt{hdf5-pure},
including the complex-typed WMP pole arrays (decoded from raw
little-endian bytes because the crate does not expose
compound types). Physics covers: energy-dependent
$\bar\nu(E)$ from prompt+delayed neutron yields (with the
per-energy interpolation of each delayed product fixed, see
Sec.~\ref{sec:offset}); tabulated outgoing energy
distributions for fission using OpenMC's stochastic bin
selection plus scaled kinematic adjustment (fixed); elastic
angular distributions from tabular $(\mu, \text{CDF})$ data
using the same stochastic-bin convention (fixed); discrete
inelastic levels MT$=51$--$91$ with real $Q$-values from HDF5
attributes; continuum inelastic (evaporation spectrum);
URR probability tables with band sampling;
S($\alpha$,$\beta$) thermal scattering for H in H$_2$O;
free-gas thermal below $400 k_B T$ with Maxwell-Boltzmann
target sampling; Shannon-entropy source-convergence monitor
on a Cartesian $8^3$ mesh; CSG geometry; PCG-$64$ PRNG; rayon
parallel transport on CPU.

The CUDA solver is event-based~\cite{tramm2024}: particles
compact by atomic alive-list scatter, energy-sorted into
$256$ log-scale bins, and processed by a persistent kernel
that runs $N$ transport steps per launch. All SVD basis is
\texttt{f64}. A runtime flag clears the uploaded pointwise
tables so SVD reconstructs every reaction channel. The WMP
evaluator lives in \texttt{src/wmp.rs} (CPU) and as a set of
\texttt{\_\_device\_\_} helpers plus an
\texttt{extern "C" \_\_global\_\_ wmp\_test\_eval} kernel in
\texttt{gpu/cuda/transport.cu} (GPU).

\paragraph{Hardware.}
We run on two systems that differ in CPU L3 capacity and GPU
class; we label them by their cache size rather than their
chassis. Small-L3 system: Intel Core i$7$-$12800$H (12th-gen
Alder Lake, $6$ P-cores + $8$ E-cores, $24$~MB L3 Smart
Cache), $32$~GB DDR$4$, NVIDIA RTX~A$1000$ ($16$ SMs,
$1.5$~MB L2). Large-L3 system: AMD Ryzen~$7$ $9800$X3D
($8$-core, $96$~MB L3 3D V-cache), $64$~GB DDR5-$6000$, NVIDIA
RTX~$3080$ ($68$~SMs, $10$~GB, $5$~MB L2). References to
``small-L3'' and ``large-L3'' below pick out the CPU+GPU pair;
the $4\times$ L3 ratio on CPU is what drives the SVD speedup
spread, and the GPU side swaps from A$1000$ to RTX~$3080$ at
the same time.

\paragraph{Reference implementation version.}
All $k_\infty$ and $k_\mathrm{eff}$ comparison numbers use
OpenMC~$0.15.3$ commit
\texttt{27e38e894697bb32a1dac7848d2618818b6b8daf}
(WSL~$2$, Ubuntu~$24.04$, conda \texttt{openmc} environment).
Nuclear data is ENDF/B-VII.$1$ (HDF5 archive from
\url{https://openmc.org/data}) for both our engine and the
OpenMC reference; cross-section evaluation uses the same
library at $T = 294$~K, with S$(\alpha,\beta)$ data
\texttt{c\_H\_in\_H2O.h5} for the PWR pin cell. Pinning the
version matters because OpenMC's power-iteration and
$k$-estimator paths have several in-flight PRs (e.g., $\#3788$,
$\#3790$, $\#3495$); between-release drift in the reference
$k$ could shift any ``$\Delta$ vs OpenMC (pcm)'' number
quoted in this paper.

\paragraph{Statistics convention.}
Per-seed $k$ estimates are averaged across active batches
with an unbiased within-run standard error; we report
$\sigma_\text{seed}$ as the sample standard deviation of the
per-seed means across $N_\text{seeds}$ seeds, and use the
standard error of the mean $\mathrm{SEM} =
\sigma_\text{seed} / \sqrt{N_\text{seeds}}$ as the noise
budget whenever a mean-to-reference comparison is quoted. A
``within $X$~SEM'' statement uses this quantity. Small-L3
sweeps use five seeds; large-L3 sweeps use ten seeds.

%% ============================================================
