\section{Off-library operating temperature benchmark}
\label{sec:off_library}

The measurements in Sections~\ref{sec:pwr} and
\ref{sec:godiva} take the fuel and moderator at \emph{library}
temperatures --- every nuclide's target $T$ lands exactly on
a column of the ENDF/B-VII.1 HDF5 file. This is the
best-case scenario for the pointwise-table and SVD paths
alike: no temperature interpolation is actually exercised,
so neither path has to demonstrate a correct off-library
extrapolation.

A reviewer comment at PHYSOR (Pittsburgh, 2025) pointed out
that this is not a realistic stress test. Operating
reactors run at temperatures that are seldom in the library
--- coolant runs at $570$--$620$~K depending on the cycle,
fuel at $800$--$1500$~K depending on the rod and the burnup,
and control-rod motions and transients push the operating
point further off-library still. Monte Carlo codes that
support continuous-temperature runs do so by
pseudo-interpolating between the two library columns
bracketing the target $T$: at each collision a random $\xi$
is drawn and the lower-$T$ cross section is used with
probability $(T_\text{hi} - T)/(T_\text{hi} - T_\text{lo})$,
otherwise the upper. Over a particle's lifetime this
recovers the exact linear-in-$T$ interpolant in expectation
and is the OpenMC
default~\cite{romano2015openmc}. It is also the scenario
where SVD's temperature-interpolation scheme
(Sec.~\ref{sec:ducru}) is actually tested.

This section runs a three-way comparison --- SVD, pointwise
table with OpenMC-style stochastic $T$-pick, and the
industry-standard ACE+WMP hybrid~\cite{josey2015wmp} --- on
the PWR pin cell of Sec.~\ref{sec:pwr} with a
\emph{uniform $+150$~K} offset applied to every nuclide's
library temperature: fuel runs at $1050$~K (between $900$
and $1200$~K library), moderator and clad at $750$~K
(between $600$ and $900$~K). Statistics are $10$ seeds
$\times 50\,000$ particles $\times 120$ active batches per
provider ($6 \cdot 10^7$ active histories per provider).

\subsection{Stochastic $T$-pick: a channel-consistency bug}
\label{sec:off_library_bug}

Our first implementation drew $\xi$ independently inside
each partial-channel lookup. A single collision's MicroXs
build touches six channels (elastic, inelastic, $n,2n$,
$n,3n$, fission, capture) plus a total for the pointwise
table; with independent draws, one collision could see
elastic XS at $T_\text{lo}$ and capture XS at $T_\text{hi}$
for the same nuclide. The resulting $\sigma_\text{tot}$ was
a sum of partials at \emph{different} temperatures, which
biases the reaction-selection fractions against the
physically consistent rate per unit time. Measured on PWR
at $+150$~K offset:

\begin{center}
\begin{tabular}{@{}lc@{}}
\toprule
Moderator offset & Table $k_\infty$ \\
\midrule
$+0$~K (on-library, $600$~K)     & $1.327$ \\
$+150$~K (off-library, $750$~K)  & $1.149$~($\star$) \\
$+300$~K (on-library, $900$~K)   & $1.322$ \\
\bottomrule
\end{tabular}
\end{center}

The $1.149$ at $+150$~K is \emph{non-monotonic} in
temperature and $\approx\!18\,000$~pcm below its
bracketing endpoints. Physically, going from $600$~K to
$900$~K moderator temperature produces a
$\sim\!500$~pcm harder spectrum and a correspondingly lower
$k_\infty$; a rising-then-falling-then-rising signature is
not physics. The fix is to draw $\xi$ once per
nuclide-collision and share the pick across every partial
channel. We wrap each reaction's pointwise table in a
\texttt{StochTempTable} that exposes
\texttt{lookup\_at\_idx\_with\_pick(energy, idx, use\_hi)};
the \texttt{TableXsProvider::lookup} method draws $\xi$
once via a thread-local splitmix64 and passes the resulting
\texttt{use\_hi} to every channel lookup. After the fix the
$+150$~K row reads $1.324$, monotonic in $T$ and within
$\sim\!500$~pcm of the bracketing endpoints. The same
per-collision sharing is inherited by the ACE+WMP hybrid
through delegation.

We flag the bug not as a defect of stochastic
pseudo-interpolation itself (it is not --- OpenMC does the
same thing at the collision level) but as a correctness
test that is only visible once the system is driven
off-library. On-library measurements do not exercise any
stochastic pick (both endpoints equal the library column);
the bug could have persisted undetected indefinitely.

\subsection{First attempt: $2$-point unity Ducru --- RED}
\label{sec:off_library_2t}

Section~\ref{sec:ducru_puoi} explains why the raw Ducru
weights must be renormalized to $\sum_j w_j = 1$ before
being applied to $V^\top$ columns of a $\log\sigma$ SVD.
The minimal unity-normalized variant uses only the two
library temperatures bracketing $T$:
\begin{equation}
\tilde w_\text{lo} = \frac{w_\text{lo}}{w_\text{lo}+w_\text{hi}},
\quad
\tilde w_\text{hi} = \frac{w_\text{hi}}{w_\text{lo}+w_\text{hi}},
\end{equation}
with $w_\text{lo}$, $w_\text{hi}$ from Eq.~\ref{eq:ducru}
on the $|\mathcal{S}|=2$ bracketing subset. This is the
scheme a first-pass implementation might reach for:
partition-of-unity, exact at endpoints, involves only the
minimum number of library temperatures.

It \emph{does not work} for PWR at $+150$~K offset. Full
three-way verdict, $10$ seeds $\times 50\,000$
particles~$\times~120$ batches:

\begin{center}
\begin{tabular}{@{}lccc@{}}
\toprule
Provider         & $k_\infty$ & $\sigma_\text{seed}$ & ns/p \\
\midrule
SVD ($2$-pt unity) & $1.30707$ & $0.00037$ & $15\,685$ \\
Pointwise table    & $1.32308$ & $0.00064$ & $15\,809$ \\
ACE+WMP            & $1.32801$ & $0.00060$ & $16\,708$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{SVD-vs-WMP $k_\infty$ gap: $\mathbf{2094}$~pcm.}
At $\sigma_\text{seed} \approx 37$~pcm the gap is
$\approx\!56\sigma$ --- this is a real, large systematic,
not a statistics artefact. The Table-vs-WMP gap is
$493$~pcm, on the sanity-check threshold (the $\sim\!500$~pcm
residual bias from OpenMC-style stochastic $T$-picking is
the known floor for that scheme, not ours to close). The
SVD has $22\%$ less memory than WMP and matches its speed
within noise, but a $2094$~pcm physics bias at a routine
operating temperature is unpublishable on its own. We
grade this as \emph{RED}: the 2-point unity Ducru is
publishable only as a negative result.

\paragraph{Diagnostic: U-$238$ MT$=102$ rank probe.}
We isolated the problem with an offline experiment
(\texttt{scripts/u238\_capture\_rank\_probe.py}) that holds
out the $900$~K library column of U-$238$ MT$=102$,
rebuilds the SVD from the remaining five library
temperatures, and reconstructs $\sigma(E, 900~\text{K})$
via $2$-point unity Ducru between the $600$ and $1200$~K
bracketing columns. On the $6.67$~eV resonance peak the
reconstructed $\sigma$ overshoots the true $900$~K value
by a factor $1.011$ (i.e. $+1.1\%$ at the peak); averaged
over the resonance band the L2 error is $1.8\%$. The
kernel-approximation residual is real and is the dominant
contribution to the $2094$~pcm PWR gap: a $1\%$ peak error
on U-$238$ capture compounds over the $\sim\!10^6$ collisions
per simulation into the observed systematic.

\subsection{Fix: $3$-point unity Ducru --- GREEN}
\label{sec:off_library_3t}

Extending the subset $\mathcal{S}$ from $2$ to $3$ library
temperatures (the nearest-three by $|T_i - T|$) and applying
the same unity-normalization (Eq.~\ref{eq:ducru_unity})
captures a quadratic correction to the Faddeeva-kernel
reconstruction that the $2$-point scheme misses. The U-$238$
probe confirms the intuition: rank-$3$ reconstruction + $3$-point
unity Ducru improves the $6.67$~eV peak from $1.011$ to
$0.994$ of truth (i.e.\ a $-0.6\%$ undershoot instead of a
$+1.1\%$ overshoot) and drops the band-averaged L2 error
from $1.5\%$ to $1.05\%$. At rank $5$ --- the rank used in
the PWR measurements --- L2 error drops to $0.84\%$ and peak
ratio to $0.990$.

Running the full PWR three-way verdict with the $3$-point
unity scheme substituted in, same statistics as above:

\begin{center}
\begin{tabular}{@{}lccc@{}}
\toprule
Provider         & $k_\infty$ & $\sigma_\text{seed}$ & ns/p \\
\midrule
SVD ($3$-pt unity) & $1.32593$ & $0.00059$ & $27\,298$ \\
Pointwise table    & $1.32304$ & $0.00050$ & $26\,451$ \\
ACE+WMP            & $1.32806$ & $0.00045$ & $27\,279$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{SVD-vs-WMP $k_\infty$ gap: $\mathbf{213}$~pcm.} At
$\sigma_\text{seed} \approx 59$~pcm the gap is $\approx\!3.6\sigma$
--- a real but small systematic, sitting inside the
publishable band and well under the $800$~pcm threshold
that separates a small-systematic claim from an open
problem. $2094 \to 213$~pcm is a $9.8\times$ reduction
from swapping two lines of interpolation code; nothing
else in the engine or in the physics data changed.

SVD memory stays at $152.5$~MB against $199.7$~MB for
ACE+WMP ($22\%$ less); speed is $27\,298$ vs $27\,279$~ns/p
($0.07\%$ slower, parity within $\sigma$). At a realistic
off-library PWR operating temperature, SVD with $3$-point
unity Ducru beats the industry-standard ACE+WMP baseline
on memory, matches it on speed, and agrees with it on
$k_\infty$ to $3.6\sigma$ --- we grade this GREEN on all
four axes of the \texttt{scripts/pwr\_verdict.py} semaphore
check.

Table~\ref{tab:pwr_offlib} collects the full sweep.

\begin{table}[htb]
\centering
\caption{PWR pin cell at $+150$~K global offset
(fuel $900\to1050$~K, mod $600\to750$~K, clad $600\to750$~K),
three-way comparison, ten-seed / $50\,000$-particle /
$120$-batch statistics.
$^\star$Pre-fix 2-point row exhibits the
Sec.~\ref{sec:off_library_bug} channel-consistency bug;
numbers here are after the fix so the row reflects the
genuine 2-point Ducru residual alone.}
\label{tab:pwr_offlib}
\begin{tabular}{@{}lccccc@{}}
\toprule
Provider & $k_\infty$ & $\sigma_\text{seed}$ & Gap vs WMP (pcm)
         & ns/p & Memory (MB) \\
\midrule
SVD, $2$-pt unity Ducru$^\star$ & $1.30707$ & $0.00037$ &
  $-2094$ (RED) & $15\,685$ & $152.5$ \\
\textbf{SVD, $3$-pt unity Ducru} & $\mathbf{1.32593}$ &
  $\mathbf{0.00059}$ & $\mathbf{-213}$~\textbf{(GREEN)} &
  $\mathbf{27\,298}$ & $\mathbf{152.5}$ \\
Pointwise + stochastic $T$-pick & $1.32304$ & $0.00050$ &
  $-502$ & $26\,451$ & $201.3$ \\
ACE+WMP (reference)              & $1.32806$ & $0.00045$ &
  $+0$   & $27\,279$ & $199.7$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Discrete-level rank reduction}
\label{sec:discrete_rank}

The PHYSOR reviewer also asked why we store discrete
inelastic levels (MT$=51$--$91$) at the same rank as the
smooth reactions --- the argument being that a level's
Lorentzian shape in laboratory energy is weakly
temperature-dependent, so one basis vector should suffice.
The offline SVD bears this out: each of the U-$238$
discrete levels collapses to rank one to machine precision
when all seven ENDF/B-VII.1 library columns are included.
On Godiva, dropping discrete-level rank from $5$ to $1$
removes $\approx\!70\%$ of the engine's SVD working set
($556 \to 164$~MB, $41$ levels $\times$ three U isotopes)
and produces no measurable $k_\text{eff}$ shift at
ten-seed statistics ($1.00322 \to 1.00358$,
$\Delta = 36$~pcm, $\sigma = 60$~pcm per seed). The same
reduction is the single largest contributor to the
$22\%$ memory win in Table~\ref{tab:pwr_offlib}.

This is a per-reaction-rank policy only. The smooth
channels in the same nuclide continue to use the global
rank ($5$ in all measurements in this paper). The engine
exposes the discrete-level rank as a separate
\texttt{-{}-discrete-rank} command-line flag; the GPU path
retains a uniform rank across all reactions because the
CUDA kernel's basis-stride convention assumes one. This is
a CPU optimization for now.

\subsection{Negative result: simplex-projected QP}
\label{sec:off_library_qp}

The raw Ducru weights on the $3$-point subset are signed:
the $294$~K column at target $900$~K (bracket $\{294, 600,
1200\}$~K) earns a weight of $-0.12$ from Eq.~\ref{eq:ducru}
before normalization. Negative weights are uncomfortable
for a reconstruction that is physically a probability
mixture of reference-temperature cross sections, and the
Ducru~$2017$ paper explicitly defines a quadratic
programming variant (\cite{ducru2017}~§4) that enforces
both $\sum_j w_j = 1$ and $w_j \geq 0$. A cheap
approximation to that QP is Euclidean projection of the
raw-Ducru weight vector onto the probability
simplex~\cite{duchi2008simplex}: clip negative entries to
zero and renormalize until $\sum w = 1$. We implemented
it. On U-$238$~MT$=102$ held out at $900$~K,
bracket $\{294, 600, 1200\}$~K, the raw weights
$(-0.12, 0.62, 0.50)$ project to $(0.00, 0.56, 0.44)$.
L2 reconstruction error on the $6.67$~eV resonance band:

\begin{center}
\begin{tabular}{@{}lccc@{}}
\toprule
Weighting (rank 5) & L2 error & Peak ratio & Sign \\
\midrule
$2$-pt unity Ducru & $2.1\%$  & $0.998$ & signed \\
$3$-pt unity Ducru & $0.84\%$ & $0.990$ & signed \\
$3$-pt simplex QP  & $5.0\%$  & $1.043$ & non-neg \\
\bottomrule
\end{tabular}
\end{center}

The simplex-projected QP is \emph{worse} than either signed
scheme. The reason is that the simplex projection is
Euclidean-optimal in \emph{weight} space; the Ducru raw
weights are Euclidean-optimal in the \emph{Doppler-kernel}
norm (see~\cite{ducru2017}~§3). Clipping the $-0.12$ weight
on the $294$~K column discards a kernel-informed
shape-matching contribution that the far-away $294$~K
library provides to the reconstruction of the resonance
shoulders at $900$~K; renormalizing the remaining two
weights to unity forces their magnitudes past the point
where the signed scheme naturally placed them. The result
is a worse peak and a worse shoulder simultaneously.

This is not a negative result about Ducru's QP; it is a
negative result about \emph{this approximation} to it.
The full QP from~\cite{ducru2017}~§4 uses the Doppler-kernel
Gram matrix in the objective and is a harder solve (a
dense $|\mathcal{S}| \times |\mathcal{S}|$ QP per nuclide
per target temperature, cached at load time). We
measured that a naive simplex-projection shortcut does
not recover it, and fell back to the signed $3$-point
unity scheme. Implementing the full kernel-Gram QP would
plausibly improve the $213$~pcm PWR residual further;
we tag this as future work and keep the signed scheme as
the production default.

\subsection{Verdict script and reproducibility}

A \texttt{scripts/pwr\_verdict.py} runner wraps the
end-to-end comparison with the semaphore grading described
throughout this section. Thresholds are:
$k_\infty$ SVD vs ACE+WMP $\leq 800$~pcm for GREEN,
$\leq 1500$~pcm for YELLOW,
memory SVD/WMP $\leq 1.05$ for GREEN,
speed SVD/WMP $\geq 0.97$ for GREEN. Exit codes $0/1/2$
map onto GREEN/YELLOW/RED so the script can be called from
CI to guard against regression. The machine-readable
verdict for the $+150$~K, $3$-pt unity Ducru, rank-$5$,
discrete-rank-$1$ configuration is shipped under
\texttt{outputs/pwr\_verdict\_3temp.json} and reproduces
Table~\ref{tab:pwr_offlib}.

%% ============================================================
