\section{Threats to validity}
\label{sec:threats}

\paragraph{Scope and nature of benchmarks.}
Two systems: Godiva (HEU-MET-FAST-$001$, three nuclides,
fast spectrum, \emph{physical} ICSBEP benchmark with measured
$k_\text{eff}^\text{exp} = 1.0000 \pm 100$~pcm $1\sigma$) and
a nine-nuclide thermal PWR pin cell (no physical benchmark;
OpenMC $0.15.3$ is used as an independent code-to-code
reference). Full-core LWR analyses carry $\gtrsim 200$
nuclides; the hybrid memory picture scales with the
reaction-channel mix and could look very different there.
On Godiva $\Delta_\text{exp}$ against ICSBEP is the
primary validation metric; $\Delta_\text{OMC}$ against
OpenMC is reported as an independent cross-code check;
the two codes straddle experiment from opposite sides at
this statistics, and the cross-code spread ($178$~pcm)
is larger than either code's offset from the measurement. On PWR,
with no physical benchmark available, the OpenMC
$|\Delta| \le 51$~pcm agreement is the strongest claim we
can make and should be read as a code-to-code cross-check,
not as absolute validation.

\paragraph{S($\alpha$,$\beta$) coverage.}
Only H in H$_2$O is exercised. D$_2$O, graphite, ZrH, and
other thermal libraries are not validated.

\paragraph{Sampling convention.}
The angular and fission outgoing-energy distributions use
stochastic-bin selection, matching OpenMC. A fully
independent reference (MCNP, Serpent) would reveal whether
OpenMC itself is biased; this is not investigated here.

\paragraph{Charged-particle absorption.}
Channels MT$=103$, $107$, first-chance fission
(MT$=19$--$21$) are absorbed into the capture macroscopic.
$k_\infty$ is correct under this convention; MT-level
reaction-rate tallies are not.

\paragraph{Statistical confidence.}
All comparisons use $\mathrm{SEM} =
\sigma_\text{seed}/\sqrt{N_\text{seeds}}$, not
$\sigma_\text{seed}$ itself. Large-L3 sweeps use
$N_\text{seeds} = 10$; hybrid sweep uses five.

\paragraph{Hybrid memory.}
The $132.9\times$ representation-byte ratio is not an
in-engine reduction. In-engine reduction is $1.06\times$
(Sec.~\ref{sec:hybrid_mem_engine}). The $132.9\times$
density ratio is reported for what it is; the
$1.06\times$ engine-level reduction is the number an
engine builder would care about.

\paragraph{GPU SVD: per-warp cache, falsified and removed.}
Earlier drafts identified a per-warp $(n, E_\text{idx})$
shared-memory cache as the engineering item that would close the
GPU SVD vs GPU pointwise gap on PWR. We implemented it, measured
it on RTX A$1000$ at five seeds $\times$ $60$ batches $\times$
$50\,000$ particles, and obtained $1.003\times$ throughput with
bit-parity within SEM --- the cache was correct but inert because
energy-bin sort granularity ($256$ bins across $17$ decades) does
not produce warp-level $(n, E_\text{idx})$ uniformity often enough
to activate the slot. The hypothesis is falsified at this
geometry. The actual fix turned out to live upstream:
\emph{synthesise MT$=4$ at HDF5 load and tabulate the per-level
CDF on a log-decimated grid}, which together remove the
$13$--$41$-deep level walk entirely (Sec.~\ref{sec:gpu}). The
cache code has been removed; the falsification is documented here.

\paragraph{GPU hybrid.}
Only the WMP evaluator itself is validated on GPU; the
transport-loop integration is not done. No in-engine GPU
hybrid $k_\infty$ or throughput is reported.

%% ============================================================
