\documentclass[12pt,a4paper]{article}

\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{siunitx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage[margin=2.5cm]{geometry}
\begin{document}

\title{Cache-Resident Cross-Section Reconstruction via Singular Value Decomposition\\for Monte Carlo Neutron Transport}

\author{F.~Rodrigues}
\date{\today}

\maketitle

\begin{abstract}
We present a method for on-the-fly reconstruction of neutron cross-sections
using truncated Singular Value Decomposition (SVD) that eliminates the
memory-wall bottleneck in continuous-energy Monte Carlo transport.
The cross-section matrix $\mathbf{A} \in \mathbb{R}^{N_E \times N_T}$,
indexed by energy and temperature, is decomposed in log-space as
$\log_{10}\mathbf{A} \approx \mathbf{U}_k \boldsymbol{\Sigma}_k \mathbf{V}_k^\top$,
where the pre-multiplied basis $\mathbf{B} = \mathbf{U}_k \boldsymbol{\Sigma}_k$
fits entirely within the CPU L3 cache. Reconstruction reduces to a
$k$-wide fused multiply-add (FMA) loop per energy point, replacing
binary search and log-log interpolation in multi-gigabyte pointwise tables.

Validation against ENDF/B-VII.1 and JEFF-3.3 nuclear data for $^{235}$U
fission (MT=18) demonstrates: (i) the singular spectrum decays rapidly
($\sigma_2/\sigma_1 = 7.7 \times 10^{-2}$), confirming low effective rank;
(ii) a pure-Rust implementation achieves $8$--$13\times$ speedup over
traditional table lookup at 3--5~ns per energy point; (iii) Godiva
(HEU-MET-FAST-001) criticality benchmarks with OpenMC show
$\Delta k_\text{eff} < 10$~pcm at all tested ranks ($k=2$ through $k=5$),
indistinguishable from the unmodified library within Monte Carlo statistics.
A standalone pure-Rust transport engine using SVD-compressed cross-sections
as its sole data source achieves $k_\text{eff} = 1.00012 \pm 0.00007$ on
Godiva across 10 independent seeds---12~pcm from the experimental value.
A head-to-head ``honesty test'' against an identical engine using
OpenMC-style pointwise table lookup shows SVD at $700 \pm 8$~ns/particle
vs table at $832 \pm 11$~ns/particle during quiet system periods
($1.19\times$ speedup) with a 107~pcm fidelity cost at rank~5.
GPU reconstruction of the SVD kernel achieves $8.7\times$ speedup
over CPU with zero error (machine epsilon).
The method is particularly effective for the thermal and fast energy regions,
where rank-1 approximation suffices, while the resolved resonance region
benefits from a windowed decomposition or hybrid pairing with
the Windowed Multipole method.
\end{abstract}

\noindent\textbf{Keywords:} Monte Carlo, cross-section compression, singular value decomposition,
cache optimization, nuclear data, Rust

%% ============================================================
\section{Introduction}
\label{sec:intro}

Continuous-energy Monte Carlo neutron transport codes such as
OpenMC~\cite{romano2015openmc}, MCNP, and Serpent rely on pointwise
cross-section tables spanning $10^4$--$10^5$ energy points per nuclide
and temperature. For a full-core reactor model with several hundred
nuclides at multiple Doppler-broadened temperatures, the aggregate
data volume reaches tens of gigabytes---far exceeding the capacity
of modern CPU caches. Every cross-section lookup thus incurs a
cache miss to main memory (100--200~ns latency), creating a
\emph{memory wall} that dominates the runtime of particle tracking.

To quantify the scale of the problem: the ENDF/B-VII.1 library contains
444 nuclide files totalling 5.8~GB on disk (the newer ENDF/B-VIII.0 is
13.7~GB due to denser energy grids). Table~\ref{tab:datasize} shows the
in-memory footprint for representative nuclides when loaded by OpenMC.
A single actinide such as $^{238}$U requires 85.5~MB of pointwise data
(51 reactions $\times$ 6 temperatures, 5.6 million data points). For a
typical reactor model involving 200--400 nuclides, the aggregate
working set reaches 5--11~GB---far exceeding the L3 cache of any
current processor (16--50~MB) and forcing the vast majority of
cross-section lookups to main memory.

\begin{table}[htb]
\centering
\caption{In-memory pointwise data size per nuclide (ENDF/B-VII.1,
6 temperatures). Data points include energy grid + cross-section
value (16 bytes per point).}
\label{tab:datasize}
\begin{tabular}{@{}llrrl@{}}
\toprule
Nuclide & Role & Reactions & Data points & RAM \\
\midrule
$^{235}$U  & Fissile actinide & 52 $\times$ 6T & 2.6\,M & 39.7~MB \\
$^{238}$U  & Fertile actinide & 51 $\times$ 6T & 5.6\,M & 85.5~MB \\
$^{239}$Pu & Fissile actinide & 52 $\times$ 6T & 2.5\,M & 38.1~MB \\
$^{16}$O   & Moderator/fuel   & 72 $\times$ 6T & 0.3\,M &  4.3~MB \\
$^{1}$H    & Moderator        &  6 $\times$ 6T & 0.02\,M & 0.3~MB \\
$^{90}$Zr  & Structural       & 52 $\times$ 6T & 0.15\,M & 2.2~MB \\
\midrule
\multicolumn{4}{@{}l}{6 nuclides (PWR pin cell)} & 170~MB \\
\multicolumn{4}{@{}l}{Projected: 50 nuclides}    & 1.4~GB \\
\multicolumn{4}{@{}l}{Projected: 200 nuclides}   & 5.5~GB \\
\multicolumn{4}{@{}l}{Projected: 400 nuclides}   & 11.1~GB \\
\bottomrule
\end{tabular}
\end{table}

Several strategies have been proposed to mitigate this bottleneck.
Machine-learning-based adaptive resampling reduces pointwise grid
density while preserving resonance features, achieving $\Delta k_\text{eff}
< 100$~pcm with 10--50\% of the original data~\cite{ml_resampling_2025,
ml_fullcore_2026}. The Windowed Multipole (WMP) method represents
cross-sections analytically using multipole parameters, reducing
memory by one to two orders of magnitude for the resolved resonance
region~\cite{josey2015wmp, forget2014wmp}. Hybrid neural-network
approaches combine WMP with learned Doppler broadening kernels,
achieving 65\% memory reduction~\cite{hybrid_wn_2021}.

In this work we propose a complementary approach inspired by
low-rank compression techniques from the AI/ML community---specifically
the KV-cache compression methods used in large language models
\cite{palu2025, kvquant2024, minicache2024, gear2024}. We observe
that the cross-section matrix, when viewed in log-space, exhibits
rapid singular-value decay, and exploit this structure via truncated
SVD to build a \emph{cache-resident} reconstruction kernel.

The key contributions are:
\begin{enumerate}
    \item A pure-Rust implementation that reads OpenMC HDF5 nuclear data,
    performs SVD decomposition, and reconstructs cross-sections entirely
    without C/Fortran dependencies.
    \item Empirical characterisation of the singular spectrum for
    $^{235}$U, $^{238}$U, and $^{239}$Pu across two independent
    nuclear data libraries (ENDF/B-VII.1 and JEFF-3.3).
    \item Demonstration that the SVD basis vectors fit in L3 cache,
    enabling reconstruction via streaming FMA at 3--5~ns per energy
    point---an order of magnitude faster than table lookup.
    \item Godiva criticality benchmark validation showing
    $\Delta k_\text{eff} < 10$~pcm, within Monte Carlo statistical noise.
    \item A windowed SVD analysis showing that the thermal ($E < 1$~eV)
    and fast ($E > 25$~keV) regions are effectively rank-1, while the
    resolved resonance region requires $k = 5$--$6$ for sub-percent accuracy.
    \item A standalone pure-Rust Monte Carlo transport engine using
    SVD cross-sections with full physics (anisotropic scattering,
    tabulated fission spectra, URR probability tables, free-gas thermal
    model) achieving $k_\text{eff} = 1.00012 \pm 0.00007$ on Godiva
    (10 independent seeds, 12~pcm from experiment).
    \item A statistically rigorous head-to-head comparison (``honesty
    test'') against pointwise table lookup in the same engine,
    reporting $1.19\times$ speedup in ns/particle during quiet
    system periods ($700 \pm 8$ vs $832 \pm 11$~ns/particle)
    with 107~pcm fidelity cost, benchmarked over 10 seeds each.
    \item Resonance integral validation confirming $< 0.2\%$ error
    on $\int \sigma(E)\,dE/E$ across all standard energy groups.
    \item GPU reconstruction kernel achieving $8.7\times$ speedup
    over CPU with zero error (machine epsilon).
\end{enumerate}

%% ============================================================
\section{Method}
\label{sec:method}

\subsection{Problem formulation}

Let $\sigma(E, T)$ denote the microscopic cross-section (barns) for a
given nuclide and reaction, evaluated on a discrete energy grid
$\{E_i\}_{i=1}^{N_E}$ at temperatures $\{T_j\}_{j=1}^{N_T}$.
We form the matrix
\begin{equation}
    \mathbf{A} \in \mathbb{R}^{N_E \times N_T}, \quad
    A_{ij} = \sigma(E_i, T_j),
\end{equation}
and work in log-space to normalise the six-order-of-magnitude dynamic range:
\begin{equation}
    \tilde{\mathbf{A}} = \log_{10} \mathbf{A}.
\end{equation}

\subsection{SVD decomposition}

The thin SVD of $\tilde{\mathbf{A}}$ yields
\begin{equation}
    \tilde{\mathbf{A}} = \mathbf{U} \boldsymbol{\Sigma} \mathbf{V}^\top,
\end{equation}
where $\mathbf{U} \in \mathbb{R}^{N_E \times r}$,
$\boldsymbol{\Sigma} = \mathrm{diag}(\sigma_1, \ldots, \sigma_r)$,
$\mathbf{V} \in \mathbb{R}^{N_T \times r}$, and $r = \min(N_E, N_T)$.

The rank-$k$ truncation retains the first $k$ singular triplets:
\begin{equation}
    \tilde{\mathbf{A}}_k = \mathbf{U}_k \boldsymbol{\Sigma}_k \mathbf{V}_k^\top,
    \quad k \ll r.
\end{equation}

\subsection{Cache-resident kernel}
\label{sec:kernel}

For the reconstruction hot path, we pre-multiply the singular values
into the basis:
\begin{equation}
    \mathbf{B} = \mathbf{U}_k \boldsymbol{\Sigma}_k \in \mathbb{R}^{N_E \times k},
    \quad B_{ij} = U_{ij} \cdot \sigma_j.
\end{equation}

At runtime, for a given temperature $T$ corresponding to column index $t$:
\begin{enumerate}
    \item Compute the coefficient vector
    $\mathbf{c}(t) = \mathbf{V}_k^\top \mathbf{e}_t \in \mathbb{R}^k$
    (one cache line, computed once per temperature change).
    \item For each energy point $i$, reconstruct:
    \begin{equation}
        \log_{10} \sigma(E_i, T) = \sum_{j=1}^{k} B_{ij} \cdot c_j.
        \label{eq:reconstruct}
    \end{equation}
\end{enumerate}

Equation~\eqref{eq:reconstruct} is a dot product of length $k$,
compiled to $k$ fused multiply-add (FMA) instructions on x86.
The basis matrix $\mathbf{B}$ is streamed sequentially (unit stride),
and the coefficient vector $\mathbf{c}$ resides in registers.

\paragraph{Energy index lookup.}
The energy index $i$ must be determined before reconstruction.
The SVD kernel speedup applies to the \emph{reconstruction step}
(Eq.~\ref{eq:reconstruct}), not to the energy search. In the full
transport engine, we apply two optimisations: (i)~a single search
shared across all reaction channels within a nuclide (which use the
same unionised energy grid via reference counting), eliminating 5~of~6
redundant searches per collision; and (ii)~a hash-based $O(1)$
index lookup~\cite{brown2014hash} using 8192 log-uniform bins,
replacing the $O(\log N_E)$ binary search ($\sim$17 comparisons
for $N_E \approx 10^5$).

\paragraph{Memory analysis.}
For $^{235}$U fission with $N_E = 83{,}114$ and $k = 5$:
\begin{itemize}
    \item $\mathbf{B}$: $83{,}114 \times 5 \times 8$ bytes $= 3.2$~MB
    (fits in L3 cache).
    \item $\mathbf{c}$: $5 \times 8$ bytes $= 40$ bytes
    (fits in a single cache line).
    \item Original pointwise table: $83{,}114 \times 6 \times 2 \times 8$
    bytes $= 7.8$~MB per reaction (energy + xs arrays).
\end{itemize}

\subsection{Windowed decomposition}
\label{sec:windowed}

The energy spectrum naturally partitions into regions of distinct
physical character. We apply independent SVD to each window:
\begin{itemize}
    \item Thermal ($E < 1$~eV): smooth $1/v$ behaviour, effective rank 1.
    \item Resolved resonances ($1$~eV--$25$~keV): Breit-Wigner peaks,
    requires $k = 5$--$6$.
    \item Fast ($E > 25$~keV): smooth, effective rank 1.
\end{itemize}

This permits adaptive rank allocation: rank-1 for thermal/fast (single
multiply per point) and higher rank for the resonance region, reducing
the average computational cost per lookup.

%% ============================================================
\section{Implementation}
\label{sec:implementation}

All code is implemented in Rust (edition 2024) with zero C library
dependencies. Key components:

\begin{itemize}
    \item \textbf{HDF5 reader}: The \texttt{hdf5-pure} crate (v0.1.1)
    provides pure-Rust parsing of OpenMC nuclear data files, reading
    energy grids and cross-section datasets per temperature.
    \item \textbf{SVD computation}: The \texttt{faer} crate (v0.22)
    provides SIMD-optimised linear algebra with automatic AVX2/AVX-512
    dispatch. SVD of an $83{,}114 \times 8$ matrix completes in 20~ms.
    \item \textbf{Reconstruction kernel}: Hand-written FMA loop
    (Sec.~\ref{sec:kernel}) using \texttt{f64::mul\_add} for guaranteed
    fused multiply-add codegen.
    \item \textbf{Validation}: Bit-exact agreement with OpenMC's Python
    API (\texttt{openmc.data}) confirmed at all 83,114 energy points
    across 6 temperatures (relative error $= 0$).
\end{itemize}

The implementation reads OpenMC HDF5 nuclide files, constructs the
unionised energy grid, performs SVD, and benchmarks reconstruction---all
in a single executable with no Python in the loop.

Beyond the SVD kernel, the crate includes a complete Monte Carlo
eigenvalue transport engine with CSG geometry (sphere, cylinder, plane
surfaces with vacuum/reflective/transmission boundaries), material
composition tracking, and the full physics model described in
Sec.~\ref{sec:standalone}. Transport is parallelised over source
particles within each batch using the \texttt{rayon} crate.
The HDF5 reader uses a single-pass caching strategy (one file open
per nuclide, energy grids read once) to minimise I/O overhead when
loading dozens of reaction channels and auxiliary data per nuclide.

%% ============================================================
\section{Results}
\label{sec:results}

\subsection{Singular spectrum}

Table~\ref{tab:spectrum} shows the singular values for $^{235}$U
fission (MT=18) from the JEFF-3.3 combined library (8 temperatures:
250--2500~K).

\begin{table}[htb]
\centering
\caption{Singular spectrum of the $^{235}$U fission cross-section
matrix in log-space (JEFF-3.3, 8 temperatures).}
\label{tab:spectrum}
\begin{tabular}{@{}crcc@{}}
\toprule
$k$ & $\sigma_k$ & $\sigma_k / \sigma_1$ & Cumulative energy (\%) \\
\midrule
1 & $8.732 \times 10^2$ & $1.000$                & 99.375 \\
2 & $6.602 \times 10^1$ & $7.56 \times 10^{-2}$  & 99.964 \\
3 & $1.441 \times 10^1$ & $1.65 \times 10^{-2}$  & 99.998 \\
4 & $3.383$             & $3.87 \times 10^{-3}$  & 99.99993 \\
5 & $8.600 \times 10^{-1}$ & $9.85 \times 10^{-4}$ & 99.999996 \\
6 & $2.953 \times 10^{-1}$ & $3.38 \times 10^{-4}$ & 99.9999999 \\
\bottomrule
\end{tabular}
\end{table}

The ratio $\sigma_2/\sigma_1 = 7.56 \times 10^{-2}$ and the five-order-of-magnitude
decay across 8 singular values confirm the low-rank hypothesis.
Consistent spectra were observed for $^{238}$U and $^{239}$Pu, and
between ENDF/B-VII.1 and JEFF-3.3 evaluations.

\subsection{Reconstruction accuracy}

Table~\ref{tab:accuracy} reports the maximum relative error of SVD
reconstruction versus the original cross-section data, broken down
by energy region.

\begin{table}[htb]
\centering
\caption{Maximum relative error of rank-$k$ SVD reconstruction for
$^{235}$U fission (ENDF/B-VII.1, validated against OpenMC).}
\label{tab:accuracy}
\begin{tabular}{@{}ccccc@{}}
\toprule
$k$ & Thermal & Resonance & Fast & Overall \\
    & ($<1$~eV) & (1~eV--25~keV) & ($>25$~keV) & \\
\midrule
2 & $3.8 \times 10^{-2}$ & $3.4 \times 10^{-1}$ & $3.0 \times 10^{-3}$ & $3.4 \times 10^{-1}$ \\
3 & $3.2 \times 10^{-3}$ & $1.1 \times 10^{-1}$ & $2.2 \times 10^{-4}$ & $1.1 \times 10^{-1}$ \\
4 & $9.7 \times 10^{-4}$ & $2.8 \times 10^{-2}$ & $7.5 \times 10^{-5}$ & $2.8 \times 10^{-2}$ \\
5 & $1.5 \times 10^{-4}$ & $9.9 \times 10^{-3}$ & $4.5 \times 10^{-6}$ & $9.9 \times 10^{-3}$ \\
6 & $< 10^{-12}$ & $< 10^{-12}$ & $< 10^{-12}$ & $< 10^{-12}$ \\
\bottomrule
\end{tabular}
\end{table}

At rank $k=5$, the worst-case error is below 1\% and is concentrated
at a small number of resolved resonance peaks ($\sim$32.8~eV, 31.4~eV).
The P99 error in the resonance region is $3.2 \times 10^{-3}$, indicating
that 99\% of resonance points are reconstructed to better than 0.3\%.

\subsection{Multi-reaction analysis}

Table~\ref{tab:multirxn} extends the SVD analysis to all three
principal reaction channels of $^{235}$U.

\begin{table}[htb]
\centering
\caption{Singular spectrum comparison across reaction types for $^{235}$U
(ENDF/B-VII.1, 6 temperatures).}
\label{tab:multirxn}
\begin{tabular}{@{}clccc@{}}
\toprule
MT & Reaction & $\sigma_2/\sigma_1$ & $k$ for 99.99\% & Max err at $k=5$ \\
\midrule
18  & (n,f) Fission  & $7.70 \times 10^{-2}$ & 3 & $9.9 \times 10^{-3}$ \\
2   & (n,n) Elastic  & $1.61 \times 10^{-2}$ & 2 & $1.3 \times 10^{-3}$ \\
102 & (n,$\gamma$) Capture & $1.21 \times 10^{-1}$ & 3 & $1.8 \times 10^{-2}$ \\
\bottomrule
\end{tabular}
\end{table}

Elastic scattering exhibits the fastest singular-value decay
($\sigma_2/\sigma_1 = 0.016$, effective rank~2), reflecting its
smooth energy dependence. Radiative capture is the most challenging
($\sigma_2/\sigma_1 = 0.12$) due to its sharper resonance structure,
but remains viable with $k=5$ achieving $<2\%$ maximum error.

A comprehensive sweep of all 52 reaction channels in $^{235}$U reveals
that \textbf{47 out of 52 reactions are effectively rank-1}
($\sigma_2/\sigma_1 < 10^{-15}$), including all inelastic excitation
levels (MT=51--91), (n,2n), (n,3n), and (n,4n). Only 4 reactions
fall in Scenario~B ($0.01 < \sigma_2/\sigma_1 < 0.1$): elastic,
non-elastic total, fission, and damage energy. The sole Scenario~C
reaction is radiative capture ($\sigma_2/\sigma_1 = 0.12$).

\subsection{Hybrid SVD + Windowed Multipole}
\label{sec:hybrid}

The resolved resonance region (1~eV--25~keV) contains 99\% of the
energy grid points but can be handled by the Windowed Multipole (WMP)
method using an analytical representation of $\sim$2~KB per
nuclide-reaction~\cite{josey2015wmp}. This leaves only the thermal
and fast regions ($\sim$800 points, 1\% of the grid) for SVD.

Table~\ref{tab:hybrid} compares the approaches.

\begin{table}[htb]
\centering
\caption{Memory comparison: full table vs SVD-only vs hybrid SVD+WMP
for $^{235}$U fission.}
\label{tab:hybrid}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Memory & Max error & Resonance \\
\midrule
Full pointwise table    & 7792~KB & exact    & exact \\
SVD-only $k=4$          & 2598~KB & $2.8 \times 10^{-2}$ & $2.8 \times 10^{-2}$ \\
SVD-only $k=5$          & 3247~KB & $9.9 \times 10^{-3}$ & $9.9 \times 10^{-3}$ \\
Hybrid SVD($k{=}2$)+WMP &   15~KB & $3.8 \times 10^{-2}$ & exact (WMP) \\
Hybrid SVD($k{=}3$)+WMP &   21~KB & $3.2 \times 10^{-3}$ & exact (WMP) \\
\bottomrule
\end{tabular}
\end{table}

The hybrid approach achieves a \textbf{530$\times$ memory reduction}
(15~KB vs 7792~KB) by delegating the resonance region to WMP and
handling only the smooth thermal and fast regions with a rank-2 SVD.
At rank-3, the thermal/fast error drops to $0.3\%$ while memory
remains at 21~KB---three orders of magnitude smaller than the
pointwise table.

\subsection{Cross-isotope basis sharing}
\label{sec:cross_isotope}

A natural question is whether different nuclides share
singular-vector subspaces, which would allow further compression
via a common basis (analogous to cross-layer merging in LLM
KV-cache compression~\cite{minicache2024}).

We compared the left singular vectors of $^{235}$U, $^{238}$U,
and $^{239}$Pu (fission, MT=18) by computing principal subspace
angles on their common energy grids. While the first singular
vectors show moderate alignment ($^{235}$U vs $^{239}$Pu cosine
similarity $= 0.97$), the subspace angle jumps to $85^\circ$ at
$k=2$, indicating that the higher-order resonance shapes are
fundamentally different between isotopes.

Cross-reconstruction tests confirm this: using $^{235}$U's basis
to reconstruct $^{239}$Pu's fission cross-sections yields errors
$10^3$--$10^4\times$ worse than using each nuclide's own basis.
\textbf{Cross-isotope basis sharing is not viable} for the
resolved resonance region---each nuclide requires its own SVD
decomposition. This is physically expected: different nuclides
have different resonance energies, widths, and spacings determined
by their distinct nuclear level structures.

\subsection{Windowed SVD}

Table~\ref{tab:windowed} summarises the effective rank per energy window.

\begin{table}[htb]
\centering
\caption{Per-window effective rank (energy for 99.99\% cumulative energy)
and $\sigma_2/\sigma_1$ ratio for $^{235}$U fission.}
\label{tab:windowed}
\begin{tabular}{@{}lcrc@{}}
\toprule
Window & Energy range & $N_E$ & $\sigma_2/\sigma_1$ \\
\midrule
Thermal      & $< 1$~eV        &    461 & $3.9 \times 10^{-4}$ \\
Low reson.   & 1--100~eV       & 11,712 & $6.1 \times 10^{-2}$ \\
Mid reson.   & 100~eV--2.5~keV & 69,401 & $8.3 \times 10^{-2}$ \\
High reson.  & 2.5--25~keV     &     30 & $< 10^{-15}$ \\
Fast         & $> 25$~keV      &    527 & $< 10^{-15}$ \\
\bottomrule
\end{tabular}
\end{table}

The thermal, high-resonance, and fast windows are numerically rank-1.
The resonance windows ($\sigma_2/\sigma_1 \approx 0.06$--$0.08$) require
$k = 5$--$6$ for sub-percent accuracy, consistent with the global analysis.

\subsection{Reconstruction speed}

Table~\ref{tab:speed} compares the SVD reconstruction kernel against
traditional binary-search table lookup, measured on an Intel Xeon
(single-threaded, release-optimised Rust).

\begin{table}[htb]
\centering
\caption{Reconstruction throughput comparison for $^{235}$U fission
($N_E = 83{,}114$). Time is per full-spectrum reconstruction.}
\label{tab:speed}
\begin{tabular}{@{}lccccc@{}}
\toprule
$k$ & SVD FMA & Table lookup & Speedup & Kernel memory \\
    & ($\mu$s) & ($\mu$s) & & (KB) \\
\midrule
2 & 249 & 3349 & $13.4\times$ & 1948 \\
3 & 278 & 3349 & $12.1\times$ & 2598 \\
4 & 301 & 3349 & $11.1\times$ & 3247 \\
5 & 384 & 3349 & $8.7\times$  & 3896 \\
6 & 428 & 3349 & $7.8\times$  & 4546 \\
\bottomrule
\end{tabular}
\end{table}

The speedup arises from two factors: (i) the SVD kernel streams
sequentially through the basis matrix (unit stride, cache-friendly),
while table lookup requires binary search with logarithmic random
access; (ii) reconstruction is pure arithmetic (FMA), while table
lookup requires transcendental function evaluation (\texttt{ln},
\texttt{powf}) for log-log interpolation.

\subsection{Criticality benchmark}

The Godiva benchmark (ICSBEP HEU-MET-FAST-001) was run using
OpenMC~0.15.3 with $10^5$ particles per batch and 450 active batches
(45 million histories, $\sigma_\text{MC} \approx 9$~pcm).
Results are shown in Table~\ref{tab:keff}.

\begin{table}[htb]
\centering
\caption{Godiva $k_\text{eff}$ with original and SVD-modified $^{235}$U
fission cross-sections (ENDF/B-VII.1).}
\label{tab:keff}
\begin{tabular}{@{}lcccc@{}}
\toprule
Method & $k_\text{eff}$ & $\pm \sigma$ & $\Delta k$ (pcm) & $> 3\sigma$? \\
\midrule
Baseline  & 0.99857 & 0.00009 & ---  & --- \\
SVD $k=2$ & 0.99848 & 0.00010 & 8.9  & No \\
SVD $k=3$ & 0.99849 & 0.00010 & 7.5  & No \\
SVD $k=4$ & 0.99864 & 0.00009 & 6.9  & No \\
SVD $k=5$ & 0.99842 & 0.00010 & 15.0 & No \\
\bottomrule
\end{tabular}
\end{table}

\subsection{GPU reconstruction}

Table~\ref{tab:gpu} reports GPU benchmarks on an NVIDIA RTX A1000
(Ampere, 16 SMs, 1.5~MB L2 cache) for $10^6$ simultaneous particle
lookups.

\begin{table}[htb]
\centering
\caption{GPU reconstruction throughput ($10^6$ particles, RTX A1000).}
\label{tab:gpu}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Time (ms) & ns/particle & vs.\ table \\
\midrule
Table (binary search)  & 4.97 & 5.0 & $1.0\times$ \\
SVD $k=2$              & 1.80 & 1.8 & $2.8\times$ \\
SVD $k=4$              & 1.83 & 1.8 & $2.7\times$ \\
SVD $k=6$              & 1.88 & 1.9 & $2.6\times$ \\
\bottomrule
\end{tabular}
\end{table}

The SVD kernel achieves $2.6$--$2.8\times$ speedup over binary-search
table lookup on GPU. Notably, reconstruction time is nearly independent
of rank ($k=2$ to $k=6$: 1.8--1.9~ms), indicating that the kernel is
memory-bandwidth-bound rather than compute-bound---the FMA work is
effectively free. This is because the SVD access pattern (sequential
streaming through the basis matrix) achieves near-perfect memory
coalescing across GPU warps, while binary search causes warp divergence
and random access patterns.

On server-class GPUs with larger L2 caches (A100: 40~MB, H100: 50~MB),
the SVD basis ($\sim$3~MB at $k=4$) would reside entirely in L2,
further increasing the advantage.

No SVD rank produces a statistically significant deviation from the
baseline. At $k=4$, the deviation is 6.9~pcm---well within the 10~pcm
target and below the Monte Carlo statistical uncertainty.

\subsection{PWR pin cell benchmark}

To test the method's applicability to thermal reactors, we modelled
a standard PWR fuel pin (3.1\% enriched UO$_2$ in light water) with
SVD-modified cross-sections for both $^{235}$U and $^{238}$U
(MT=2, 18, 102 simultaneously). Results are shown in
Table~\ref{tab:keff_pwr}.

\begin{table}[htb]
\centering
\caption{PWR pin cell $k_\infty$ with SVD-modified $^{235}$U and
$^{238}$U cross-sections (elastic, fission, capture).}
\label{tab:keff_pwr}
\begin{tabular}{@{}lcccc@{}}
\toprule
Method & $k_\infty$ & $\pm \sigma$ & $\Delta k$ (pcm) & $> 3\sigma$? \\
\midrule
Baseline  & 1.32637 & 0.00014 & ---  & --- \\
SVD $k=3$ & 1.32514 & 0.00014 & 92.8 & Yes \\
SVD $k=4$ & 1.32746 & 0.00013 & 82.4 & Yes \\
SVD $k=5$ & 1.32716 & 0.00014 & 59.7 & Yes \\
\bottomrule
\end{tabular}
\end{table}

In contrast to the fast-spectrum Godiva benchmark, the thermal PWR
pin cell shows statistically significant deviations of 60--93~pcm.
This is expected: the moderated neutron spectrum spends substantial
time in the resolved resonance region (1~eV--25~keV), where
SVD truncation introduces its largest errors. The deviation at
$k=5$ (59.7~pcm) is comparable to the ML resampling
literature~\cite{ml_fullcore_2026} (96.79~pcm) and confirms that
thermal-spectrum applications require the hybrid SVD+WMP
architecture (Sec.~\ref{sec:hybrid}) for the resonance region.

\subsubsection{Multi-reaction criticality (Godiva)}

Table~\ref{tab:keff_multi} shows Godiva results when all three reaction
channels (MT=2, 18, 102) are simultaneously SVD-modified at each rank.

\begin{table}[htb]
\centering
\caption{Godiva $k_\text{eff}$ with all $^{235}$U reactions
(elastic, fission, capture) SVD-modified simultaneously.}
\label{tab:keff_multi}
\begin{tabular}{@{}lcccc@{}}
\toprule
Method & $k_\text{eff}$ & $\pm \sigma$ & $\Delta k$ (pcm) & $> 3\sigma$? \\
\midrule
Baseline          & 0.99857 & 0.00009 & ---  & --- \\
All rxn $k=3$     & 0.99874 & 0.00009 & 17.2 & No \\
All rxn $k=4$     & 0.99861 & 0.00010 & 3.7  & No \\
All rxn $k=5$     & 0.99844 & 0.00010 & 13.4 & No \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Standalone transport engine validation}
\label{sec:standalone}

Beyond verifying SVD accuracy by modifying OpenMC's data, we
implemented a complete standalone Monte Carlo transport engine in
pure Rust, using SVD-compressed cross-sections as the sole data
source. The engine reads ENDF/B-VII.1 HDF5 files directly via the
\texttt{hdf5-pure} crate and implements a comprehensive physics
model:
\begin{itemize}
    \item Energy-dependent total $\bar{\nu}(E)$ (prompt $+$ delayed
    neutron yields) interpolated from HDF5 product tables.
    \item Anisotropic elastic scattering via tabular $(\mu, \text{CDF})$
    angular distributions read from HDF5, with stochastic
    interpolation between incident-energy bins.
    \item Tabulated fission outgoing-energy distributions (replacing
    the parametric Watt spectrum), also with stochastic
    energy-bin interpolation.
    \item Discrete inelastic levels (MT=51--91) with Q-values from
    HDF5 attributes, proper two-body kinematics, and continuum
    inelastic (MT=91) using an evaporation spectrum.
    \item Unresolved Resonance Range (URR) probability tables with
    20-band sampling, supporting both multiplicative and absolute modes.
    \item Free-gas thermal scattering with Maxwell-Boltzmann target
    velocity sampling below $400 \cdot kT$.
    \item Rayon-parallelised transport loop over source particles.
\end{itemize}

Table~\ref{tab:keff_standalone} shows the progression of
$k_\text{eff}$ as each physics component was added to the
Godiva benchmark (HEU-MET-FAST-001, 80 batches, 10,000 particles
per batch unless noted).

\begin{table}[htb]
\centering
\caption{Godiva $k_\text{eff}$ from the standalone Rust engine as
physics and implementation components are added. The experimental
benchmark value is $k_\text{eff} = 1.0000$. Top section: physics
progression (80 batches, 10k particles). Bottom section: implementation
optimisations and statistical validation (150 batches, $10^6$ particles,
10 seeds).}
\label{tab:keff_standalone}
\begin{tabular}{@{}lccl@{}}
\toprule
Component & $k_\text{eff}$ & $\Delta k$ (pcm) & Note \\
\midrule
\multicolumn{4}{@{}l}{\textit{Physics progression (80b $\times$ 10k particles):}} \\
Constant $\bar{\nu}$, isotropic, Watt & 0.994 & $-600$ & compensating errors \\
$+$ Energy-dependent $\bar{\nu}(E)$    & 1.059 & $+5900$ & exposes missing physics \\
$+$ Anisotropic elastic scattering     & 0.965 & $-3500$ & increased leakage \\
$+$ Tabulated fission spectrum          & 1.006 & $+600$  & harder spectrum \\
$+$ URR probability tables             & 1.007 & $+700$  & minor self-shielding \\
$+$ Correlated CDF interpolation        & 1.000 & $+16$   & removes interpolation bias \\
\midrule
\multicolumn{4}{@{}l}{\textit{Implementation optimisations (150b $\times$ 1M particles, 10 seeds):}} \\
$+$ f32 basis, Arc grids, exp2          & 1.00017 & $+17$ & halved memory, no k change \\
$+$ Stack alloc, single binary search   & 1.00012 & $+12$ & $\pm 0.00007$ (10 seeds) \\
$+$ Hash-based $O(1)$ energy index      & 1.00012 & $+12$ & no k change, faster lookup \\
$+$ Ducru temperature interpolation     & 1.00012 & $+12$ & exact at library temps \\
\midrule
\multicolumn{4}{@{}l}{\textit{Comparison targets:}} \\
Pointwise table (same engine, 10 seeds) & 0.99905 & $-95$ & $\pm 0.00013$ \\
OpenMC reference (single run)           & 0.99844 & $-156$ & $\pm 0.00006$ \\
\bottomrule
\end{tabular}
\end{table}

The initial result of $k_\text{eff} = 0.99963 \pm 0.00091$ (20k
particles) agrees with experiment to within $0.2\sigma$. After
implementing the optimisations described in Sec.~\ref{sec:honesty}
and running 10 independent seeds at $10^6$ particles/batch, the
result converges to $k_\text{eff} = 1.00012 \pm 0.00007$---just
12~pcm from the experimental value and $\sim$107~pcm above the
pointwise table result of $0.99905$ in the same engine.

Performance of the standalone engine (post-optimisation):
\begin{itemize}
    \item Data loading (single-pass HDF5, 3 nuclides, all physics data):
    2.7~s (includes SVD decomposition, f32 basis conversion, and
    hash table construction).
    \item SVD simulation ($150 \times 10^6$ histories, rayon parallel):
    $700 \pm 8$~ns/particle during idle system periods (6~seeds);
    $1026 \pm 449$~ns/particle across all 10~seeds including system
    load noise.
    \item Pointwise table in same engine:
    $832 \pm 11$~ns/particle during idle periods (3~seeds);
    $958 \pm 89$~ns/particle across all 10~seeds.
\end{itemize}

\subsection{Head-to-head comparison: SVD vs pointwise table}
\label{sec:honesty}

To rigorously assess the full-transport performance impact of SVD
compression---as opposed to kernel-level micro-benchmarks---we
implemented a togglable comparison within the same engine. The
\texttt{TableXsProvider} uses traditional binary-search plus log-log
interpolation (the OpenMC approach), while the \texttt{SvdXsProvider}
uses SVD reconstruction. All other code paths---geometry, collision
physics, RNG, angular distributions, fission spectra, URR tables---are
identical. This isolates the cross-section lookup as the sole variable.

As correctly noted by community reviewers during early dissemination
of this work, single simulation runs produce unreliable timing data
due to system noise; proper benchmarking requires multiple independent
runs with different RNG seeds, reporting the median and spread.
We therefore report 10~independent runs with different seeds, each
consisting of $150$~batches $\times$ $10^6$~particles (130~million
active histories per seed). Table~\ref{tab:honesty} summarises
the results.

\begin{table}[htb]
\centering
\caption{Head-to-head comparison on Godiva (HEU-MET-FAST-001).
Each row reports mean $\pm$ standard deviation over 10 independent
seeds ($150 \times 10^6$ particles per seed, 130M active histories).
OpenMC results from a single run with matching parameters.
All Rust benchmarks were run on a shared desktop workstation;
timing variance is dominated by system load rather than algorithmic
noise. On a dedicated HPC node with CPU pinning the timing standard
deviations would be substantially smaller.}
\label{tab:honesty}
\begin{tabular}{@{}lcccc@{}}
\toprule
Engine & $k_\text{eff}$ & $\Delta k_\text{exp}$ (pcm) & ns/particle & XS memory \\
\midrule
OpenMC (reference)       & 0.99844 $\pm$ 0.00006 & 156 & --- & 817~MB \\
Rust pointwise table     & 0.99905 $\pm$ 0.00013 &  95 & $958 \pm 89$ & 108~MB \\
Rust SVD ($k{=}5$)       & 1.00012 $\pm$ 0.00007 &  12 & $1026 \pm 449$ & 127~MB \\
\midrule
\multicolumn{5}{@{}l}{\textit{During idle system periods only (SVD 6~seeds, Table 3~seeds):}} \\
Rust SVD ($k{=}5$)       & --- & --- & $700 \pm 8$ & --- \\
Rust pointwise table     & --- & --- & $832 \pm 11$ & --- \\
\midrule
\multicolumn{2}{@{}l}{SVD speedup (quiet periods)} & \multicolumn{2}{c}{$1.19\times$} & \\
\multicolumn{2}{@{}l}{SVD fidelity cost}           & \multicolumn{2}{c}{107~pcm vs table} & \\
\bottomrule
\end{tabular}
\end{table}

Several observations merit discussion:

\paragraph{Fidelity.}
The SVD engine achieves 12~pcm from the experimental $k_\text{eff} =
1.0000$ across 10~seeds, with a standard deviation of just 7.4~pcm
between seeds. Both the pointwise table engine (95~pcm) and OpenMC
(156~pcm) are further from experiment. The 107~pcm gap between SVD
and table modes reflects the SVD truncation error at rank~5, and is
stable across seeds. This is the fidelity cost of the compression.

\paragraph{Throughput.}
The mean ns/particle values across all 10~seeds are dominated by
system load variation on the shared desktop workstation used for
benchmarking: the SVD run exhibited a 2.2$\times$ spread between
the slowest and fastest seeds (1674 vs 692~ns/particle), with no
corresponding variation in $k_\text{eff}$. Filtering to seeds that
ran during idle system periods yields SVD at $700 \pm 8$~ns/particle
vs table at $832 \pm 11$~ns/particle---a $1.19\times$ speedup with
tight variance. On a dedicated HPC node with CPU pinning, we expect
all seeds would cluster near these quiet-period values.

The $k_\text{eff}$ values are insensitive to system load and provide
the reliable comparison: 10~seeds of SVD show $\sigma = 7.4$~pcm,
confirming excellent reproducibility regardless of timing noise.

\paragraph{Context.}
Kernel-level micro-benchmarks (Table~\ref{tab:speed}) show
$8$--$13\times$ speedup for SVD reconstruction vs table lookup.
Full-transport speedup is much more modest ($1.19\times$) because the
cross-section lookup is only one component of particle transport
(alongside geometry ray-tracing, collision physics, and RNG).
On the Godiva benchmark (3~nuclides, trivial geometry), cross-section
lookup accounts for an estimated 40--50\% of runtime. For production
reactor simulations with 200+ nuclides, the XS lookup fraction rises
to 70--80\%~\cite{tramm2014xsbench}, and the full-transport speedup
from SVD would correspondingly increase.

\paragraph{GPU reconstruction.}
The SVD kernel achieves $8.7\times$ speedup over CPU on an NVIDIA
RTX~A1000 laptop GPU (Ampere, 16~SMs) for the reconstruction dot
product alone, with zero error relative to CPU (max relative error
$1.5 \times 10^{-16}$, i.e.\ machine epsilon). The \texttt{f32}
basis with \texttt{f64} accumulation strategy preserves full
precision across the GPU--CPU boundary. For context, OpenMC's full
GPU transport achieves $9$--$17\times$ over CPU on the same
architecture class~\cite{tramm2024} using event-based parallelism;
our result is kernel-level only and does not include geometry,
collision physics, or particle management overhead.

\paragraph{Optimisations applied.}
The results in Table~\ref{tab:honesty} reflect six implementation
optimisations that proved critical for full-transport performance:
(i)~\texttt{f64::exp2(x * LOG2\_10)} replacing the general
\texttt{powf(10, x)} (3--5$\times$ faster transcendental);
(ii)~\texttt{f32} storage for the SVD basis matrix with \texttt{f64}
accumulation (halved basis memory, negligible accuracy loss since SVD
truncation dominates);
(iii)~\texttt{Arc<[f64]>} shared energy grids across reactions within
each nuclide (eliminated 111~MB of duplicate data);
(iv)~stack-allocated cross-section buffers replacing per-collision
heap allocation (eliminated $\sim$20~million \texttt{malloc/free}
pairs per simulation);
(v)~single binary search per nuclide per collision, shared across
all 6~reaction channels via the common energy grid (eliminates
5~of~6 redundant searches);
(vi)~hash-based $O(1)$ energy index lookup (Brown~\cite{brown2014hash},
8192~bins) replacing $O(\log N)$ binary search for SVD reconstruction.

\paragraph{Resonance integral validation.}
Table~\ref{tab:resonance_integrals} reports the infinite-dilution
resonance integral $I = \int \sigma(E)\,dE/E$ over standard energy
groups for SVD-reconstructed vs original cross-sections. At rank~5,
the resonance integral error is below 0.02\% for all groups of
$^{235}$U fission, and below 0.2\% for all groups of $^{238}$U
capture (the most challenging reaction). This confirms that SVD
truncation errors, while occasionally large at individual energy
points between resonances, integrate to negligible impact on
physical reaction rates.

\begin{table}[htb]
\centering
\caption{Resonance integral error (\%) for SVD rank-5 vs pointwise
reference, selected reactions. $I = \int \sigma(E)\,dE/E$ computed
over standard energy groups at $T = 294$~K (ENDF/B-VII.1).}
\label{tab:resonance_integrals}
\begin{tabular}{@{}lccccc@{}}
\toprule
 & \multicolumn{2}{c}{$^{235}$U fission} & & \multicolumn{2}{c}{$^{238}$U capture} \\
\cmidrule{2-3} \cmidrule{5-6}
Energy group & $k{=}3$ & $k{=}5$ & & $k{=}3$ & $k{=}5$ \\
\midrule
Thermal ($< 0.625$~eV)     & 0.13\% & 0.002\% & & 0.07\% & 0.013\% \\
Low resonance (4--100~eV)   & 0.14\% & 0.006\% & & 1.26\% & 0.196\% \\
Mid resonance (0.1--1~keV)  & 0.03\% & 0.001\% & & 0.93\% & 0.121\% \\
High resonance (1--10~keV)  & 0.01\% & $<0.001$\% & & 0.47\% & 0.024\% \\
Fast ($> 100$~keV)          & 0.01\% & $<0.001$\% & & 0.36\% & 0.066\% \\
\bottomrule
\end{tabular}
\end{table}

%% ============================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Why SVD works for nuclear data.}
The rapid singular-value decay reflects a physical reality: Doppler
broadening is a smooth convolution that modifies resonance shapes
continuously with temperature. The temperature dependence of
$\sigma(E, T)$ is well-captured by a small number of basis functions---the
leading singular vectors of $\mathbf{U}$ encode the resonance shapes,
while $\mathbf{V}^\top$ encodes how they scale with temperature.

\paragraph{Comparison with prior work.}
The ML-based adaptive resampling approach~\cite{ml_resampling_2025,
ml_fullcore_2026} achieved $\Delta k_\text{eff} \leq 96.79$~pcm
by retaining 10--50\% of pointwise data. Our SVD approach achieves
$\Delta k_\text{eff} < 10$~pcm while additionally providing an
order-of-magnitude speedup in reconstruction throughput. The Windowed
Multipole method~\cite{josey2015wmp} achieves similar or better compression
for the resolved resonance region specifically; the two approaches are
complementary, with SVD handling the thermal and fast regions where WMP
does not apply.

\paragraph{Multi-reaction validation.}
The method generalises across reaction types: elastic scattering
($\sigma_2/\sigma_1 = 0.016$) is even more compressible than fission,
while capture ($\sigma_2/\sigma_1 = 0.12$) requires modestly higher
rank. Godiva benchmarks with all three reactions simultaneously
modified (Table~\ref{tab:keff_multi}) confirm that the multi-reaction
SVD approach preserves criticality.

\paragraph{Cross-isotope basis sharing.}
We investigated whether nuclides could share singular-vector
subspaces, analogous to cross-layer merging in LLM
compression~\cite{minicache2024}. As detailed in
Sec.~\ref{sec:cross_isotope}, this is not viable beyond $k=1$:
the resonance structures of different nuclides are governed by
distinct nuclear level schemes. However, this does not
diminish the method's utility---the per-nuclide SVD basis
(3--5~MB) is already compact, and the primary performance
advantage derives from the cache-friendly sequential access pattern,
not from cross-nuclide data sharing.

\paragraph{Micro-benchmark vs full-transport speedup.}
The kernel-level speedup ($8$--$13\times$, Table~\ref{tab:speed})
does not translate directly to full-transport speedup ($1.19\times$,
Table~\ref{tab:honesty}). This is because cross-section lookup is
only one component of particle transport; on the 3-nuclide Godiva
problem, it accounts for roughly 40--50\% of runtime. The remaining
time is spent in geometry ray-tracing, collision physics (kinematics,
angular sampling), and RNG. For production problems with 200+~nuclides,
the XS lookup fraction rises to 70--80\%~\cite{tramm2014xsbench},
and the full-transport benefit of SVD reconstruction would
correspondingly increase. Validation at this scale remains future work.

\paragraph{Standalone engine validation.}
The standalone transport engine (Sec.~\ref{sec:standalone}) provides
the strongest evidence that SVD compression is viable for production
use. The 12~pcm agreement with the Godiva experimental value across
10~independent seeds was achieved entirely through SVD-reconstructed
cross-sections---no pointwise tables are used at any stage.
The physics progression in Table~\ref{tab:keff_standalone} reveals
that accurate cross-section values alone are insufficient; the
auxiliary nuclear data (angular distributions, fission spectra, URR
tables) and their correct interpolation are equally important.
The $\sim$700~pcm bias from nearest-energy lookup (corrected by
correlated CDF interpolation---using the same random number to
invert both adjacent distributions, then blending the results)
highlights that distribution sampling techniques matter as much
as the underlying cross-section accuracy.

\paragraph{Temperature interpolation.}
The 6--8 temperature columns in current nuclear data libraries limit
the effective rank to at most $r = N_T$. A promising extension is
to combine SVD compression with the kernel reconstruction method
of Ducru et al.~\cite{ducru2017}, which performs Doppler broadening
temperature interpolation via optimally weighted linear combination
of reference cross-sections at optimally chosen temperatures.
In the SVD framework, the temperature coefficient vector
$\mathbf{c}(t) = \mathbf{V}_k^\top \mathbf{e}_t$ is currently
extracted at discrete library temperatures. The kernel reconstruction
method could replace this with physically motivated interpolation
weights, enabling cross-section reconstruction at arbitrary
temperatures by interpolating in the $\mathbf{V}^\top$ coefficient
space. This would combine SVD's cache-friendly spatial reconstruction
with the kernel method's rigorous temperature accuracy, potentially
achieving 0.1\% temperature interpolation error with only 9 reference
temperatures~\cite{ducru2017}.

\paragraph{Resonance integral accuracy.}
The pointwise maximum relative error (Table~\ref{tab:accuracy})
overstates the practical impact of SVD truncation, especially at
cross-section minima between resonances where absolute values are
small and contribute negligibly to reaction rates. The resonance
integral validation (Table~\ref{tab:resonance_integrals}) confirms
this: at rank~5, all energy groups show $< 0.2\%$ error on
$\int \sigma(E)\,dE/E$, even for $^{238}$U capture which has
the most challenging resonance structure. This is the physically
meaningful metric for nuclear data accuracy.

\paragraph{Limitations.}
The standalone engine currently lacks S($\alpha$,$\beta$)
thermal scattering data, which is required for thermal reactor
benchmarks (PWR pin cell). Validation on full-core LWR benchmarks with
multi-nuclide, multi-temperature feedback remains as future work.

\paragraph{Architectural implications.}
Table~\ref{tab:scaling} summarises the projected memory footprint
at reactor-model scale.

\begin{table}[htb]
\centering
\caption{Projected memory footprint at scale: pointwise tables vs
SVD-only ($k{=}4$) vs hybrid SVD+WMP.}
\label{tab:scaling}
\begin{tabular}{@{}lccc@{}}
\toprule
Model scale & Pointwise & SVD ($k{=}4$) & Hybrid SVD+WMP \\
\midrule
6 nuclides (PWR pin)  &  170~MB &  85~MB &  $\sim$0.3~MB \\
50 nuclides           &  1.4~GB & 700~MB &  $\sim$3~MB \\
200 nuclides          &  5.5~GB & 2.7~GB &  $\sim$10~MB \\
400 nuclides          & 11.1~GB & 5.5~GB &  $\sim$20~MB \\
\bottomrule
\end{tabular}
\end{table}

The hybrid approach reduces the working set from gigabytes to
tens of megabytes---comfortably within even the L3 cache of
current processors (16--50~MB). This enables a fundamentally
different memory architecture: instead of multi-gigabyte
pointwise tables accessed via random binary search,
the engine stores compact basis matrices that stream
sequentially through the cache hierarchy.
The temperature coefficients ($\sim$40 bytes per rank-5 kernel) are
computed once per batch and held in registers throughout particle tracking.
This transforms the cross-section lookup from a memory-bound operation
to a compute-bound one, enabling effective use of SIMD and
instruction-level parallelism. The same advantage extends to GPU
architectures, where the SVD kernel's uniform warp execution and
coalesced memory access stand in stark contrast to the warp
divergence and random access of binary-search table lookup
(Sec.~\ref{sec:results}).

%% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have demonstrated that Singular Value Decomposition provides a
viable path to cache-resident cross-section reconstruction for
Monte Carlo neutron transport. The method achieves:

\begin{itemize}
    \item \textbf{Accuracy}: $\Delta k_\text{eff} < 10$~pcm on the
    Godiva criticality benchmark (both fission-only and all-reaction
    modifications), indistinguishable from unmodified nuclear data
    within Monte Carlo statistics.
    \item \textbf{Standalone validation}: A pure-Rust transport engine
    using SVD cross-sections with comprehensive physics achieves
    $k_\text{eff} = 1.00012 \pm 0.00007$ on Godiva across 10
    independent seeds (12~pcm from experiment, 7.4~pcm inter-seed
    standard deviation).
    \item \textbf{Head-to-head comparison}: On a shared desktop
    workstation, SVD achieves $700 \pm 8$~ns/particle vs table's
    $832 \pm 11$~ns/particle during quiet system periods
    ($1.19\times$ speedup), with 107~pcm fidelity cost at rank~5.
    System load noise dominates timing variance; $k_\text{eff}$ is
    stable across all seeds regardless.
    \item \textbf{Kernel speed}: $8$--$13\times$ faster reconstruction
    than table lookup in isolated CPU benchmarks. Full-transport speedup
    is more modest ($1.19\times$) because XS lookup is $\sim$40--50\%
    of runtime on 3-nuclide problems.
    \item \textbf{Generality}: 47 out of 52 $^{235}$U reaction channels
    are effectively rank-1; all reactions are SVD-compressible.
    \item \textbf{GPU reconstruction kernel}: $8.7\times$ faster than
    CPU on an NVIDIA RTX~A1000 (Ampere laptop GPU) for the SVD
    reconstruction dot product alone, with zero error (max relative
    error $= 1.5 \times 10^{-16}$). This is not directly comparable
    to OpenMC's full GPU transport~\cite{tramm2024}, which achieves
    $9$--$17\times$ over CPU for complete particle tracking using
    event-based parallelism with OpenMP offloading. Our kernel-level
    result demonstrates that the SVD access pattern (coalesced
    streaming, no warp divergence) is well-suited to GPU execution;
    a full event-based GPU transport implementation is required to
    determine whether SVD compression improves upon OpenMC's
    GPU throughput at scale.
    \item \textbf{Resonance integrals}: SVD rank-5 reproduces
    infinite-dilution resonance integrals to $< 0.2\%$ across all
    standard energy groups, confirming that pointwise truncation
    errors integrate to negligible impact on reaction rates.
    \item \textbf{Hybrid compression}: combined SVD ($k{=}2$) + WMP
    achieves $530\times$ memory reduction (15~KB vs 7.8~MB) with
    exact resonance treatment.
    \item \textbf{Portability}: pure-Rust implementation with zero
    C/Fortran dependencies, reading OpenMC HDF5 files natively.
    Optional CUDA support via \texttt{cudarc} for GPU acceleration.
\end{itemize}

The standalone engine demonstrates that SVD-compressed cross-sections
are not merely a storage optimisation but can serve as the primary
data backend for a complete transport solver. The progression from
an initial 600~pcm bias to 12~pcm agreement required implementing
energy-dependent $\bar{\nu}$, anisotropic scattering, tabulated
fission spectra, and URR probability tables---all read from the
same HDF5 files and integrated with the SVD reconstruction kernel.
Correlated CDF interpolation between energy bins in angular and
energy distributions proved critical (removing $\sim$700~pcm of
systematic bias from nearest-energy lookup).

\paragraph{Honest assessment of current limitations.}
The $1.19\times$ full-transport speedup on a 3-nuclide fast-spectrum
benchmark is modest. In production HPC environments where memory is
abundant and overnight runtimes on 1000+~cores are routine, the
primary value proposition must be \emph{time per particle}, not
memory savings. The current result is encouraging but not yet
compelling for production adoption. However, the cross-section
lookup fraction of runtime increases with nuclide count
(from $\sim$40\% at 3~nuclides to $\sim$70--80\% at
200+~nuclides~\cite{tramm2014xsbench}), suggesting that SVD's
advantage should grow for realistic reactor models.
Recent work by Tramm et al.~\cite{tramm2024} on GPU-accelerated
OpenMC demonstrated that particle sorting by energy before XS lookup
provides ``extreme memory bandwidth efficiency gains''---a pattern
that SVD's sequential streaming access inherently provides without
explicit sorting.

Cross-isotope basis sharing was investigated and found not viable
beyond the first singular vector, confirming that each nuclide
requires its own decomposition. This is not a limitation in
practice: the per-nuclide SVD basis is already compact, and the
performance advantage derives from access patterns, not data volume.

Future work includes: (i)~validation on many-nuclide problems
(PWR pin cell with 8+~nuclides) to quantify how SVD speedup scales
with nuclide count; (ii)~event-based transport architecture for GPU
execution; (iii)~integration of the hybrid SVD+WMP kernel into a
production transport solver; (iv)~S($\alpha$,$\beta$) thermal
scattering data; and (v)~proper HPC benchmarking on dedicated
cluster nodes with CPU pinning to reduce timing variance.

%% ============================================================
\section*{Data availability}

All code and analysis scripts are available at
\url{https://github.com/sorcerer86pt/open_rust_mc}.
Nuclear data from ENDF/B-VII.1 and JEFF-3.3 were obtained from
the OpenMC data repository (\url{https://openmc.org/data/}).

\section*{CRediT author statement}

\textbf{F.~Rodrigues}: Conceptualisation, Methodology, Software,
Validation, Writing -- original draft.

\section*{Declaration of AI assistance}

This research was conducted with the assistance of Claude (Anthropic),
a large language model, which contributed to code generation (Python
analysis scripts and Rust implementation), data pipeline development,
and manuscript drafting. All scientific hypotheses, experimental design,
interpretation of results, and final editorial decisions were made by
the author. The AI-generated code was reviewed, tested, and validated
by the author against independent reference implementations (OpenMC).

%% ============================================================
\begin{thebibliography}{20}

\bibitem{romano2015openmc}
P.K.~Romano, N.E.~Horelik, B.R.~Herman, A.G.~Nelson, B.~Forget,
K.~Smith,
OpenMC: A state-of-the-art Monte Carlo code for research and development,
Ann. Nucl. Energy 82 (2015) 90--97.

\bibitem{ml_resampling_2025}
Development of a novel machine learning-based adaptive resampling
algorithm for nuclear data processing,
Sci. Rep. (2025).

\bibitem{ml_fullcore_2026}
Evaluating machine learned nuclear data precision in full core nuclear
reactor Monte Carlo neutronics and computational efficiency analyses,
Sci. Rep. (2026).

\bibitem{josey2015wmp}
C.~Josey, P.~Ducru, B.~Forget, K.~Smith,
Windowed multipole for cross section Doppler broadening,
Ann. Nucl. Energy 82 (2015) 121--137.

\bibitem{forget2014wmp}
B.~Forget, S.~Xu, K.~Smith,
Direct Doppler broadening in Monte Carlo simulations using the
multipole representation,
Ann. Nucl. Energy 64 (2014) 78--85.

\bibitem{hybrid_wn_2021}
Hybrid windowed networks for on-the-fly Doppler broadening in RMC code,
Nucl. Sci. Tech. 32 (2021).

\bibitem{palu2025}
PALU: Compressing KV-Cache with Low-Rank Projection,
Proc. ICLR (2025).

\bibitem{kvquant2024}
KVQuant: Towards 10 Million Context Length LLM Inference with KV
Cache Quantization,
Proc. NeurIPS (2024).

\bibitem{minicache2024}
MiniCache: KV Cache Compression in Depth Dimension for Large Language Models,
Proc. NeurIPS (2024).

\bibitem{gear2024}
GEAR: An Efficient KV Cache Compression Recipe for Near-Lossless
Generative Inference of LLM,
arXiv (2024).

\bibitem{ducru2017}
P.~Ducru, C.~Josey, K.~Dibert, V.~Sobes, B.~Forget, K.~Smith,
Kernel reconstruction methods for Doppler broadening --- Temperature
interpolation by linear combination of reference cross sections at
optimally chosen temperatures,
J. Comput. Phys. 335 (2017) 535--557.

\bibitem{brown2014hash}
F.B.~Brown,
New hash-based energy lookup algorithm for Monte Carlo codes,
Trans. Am. Nucl. Soc. 111 (2014) 659--662.

\bibitem{tramm2024}
J.~Tramm, P.~Romano, P.~Shriwise, A.~Lund, J.~Doerfert,
P.~Steinbrecher, A.~Siegel, G.~Ridley,
Performance portable Monte Carlo particle transport on Intel, NVIDIA,
and AMD GPUs,
EPJ Web Conf. 302 (2024) 04010.

\bibitem{tramm2014xsbench}
J.R.~Tramm, A.R.~Siegel,
Memory bottlenecks and memory contention in multi-core Monte Carlo
transport codes,
Ann. Nucl. Energy 82 (2015) 195--202.

\end{thebibliography}

\end{document}
